---
title: 'Lab 5: Machine Learning in Hydrology'
format:
   html:
    toc: true
subtitle: 'Using Tidymodels & CAMELS Data'
author:
  - name: Taia Storrs
    email: taia.storrs@colostate.edu
---

# Introduction

In this lab, we will explore predictive modeling in hydrology using the `tidymodels` framework and the CAMELS (Catchment Attributes and Meteorology for Large-sample Studies) dataset.

```{r}
#| message: false
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
library(ranger)      
library(xgboost)     
library(sf)          
library(spatialsample) 
library(skimr)       
library(visdat)     
library(patchwork)   
```


## What is tidymodels?

`tidymodels` is an R framework designed for machine learning and statistical modeling. Built on the principles of the `tidyverse`, `tidymodels` provides a consistent and modular approach to tasks like data preprocessing, model training, evaluation, and validation. By leveraging the strengths of packages such as `recipes`, `parsnip`, and `yardstick`, `tidymodels` streamlines the modeling workflow, making it easier to experiment with different models while maintaining reproducibility and interpretability.

## What is the CAMELS dataset?

The CAMELS dataset is a widely used resource in hydrology and environmental science, providing data on over 500 self-draining river basins across the United States. It includes meteorological forcings, streamflow observations, and catchment attributes such as land cover, topography, and soil properties. This dataset is particularly valuable for large-sample hydrology studies, enabling researchers to develop and test models across diverse climatic and physiographic conditions.

In this lab, we will focus on predicting mean streamflow for these basins using their associated characteristics. CAMELS has been instrumental in various hydrologic and machine learning applications, including:

 - Calibrating Hydrologic Models – Used for parameter tuning in models like SAC-SMA, VIC, and HBV, improving regional and large-sample studies.

 - Training Machine Learning Models – Supports deep learning (e.g., LSTMs) and regression-based streamflow predictions, often outperforming traditional methods.

 - Understanding Model Behavior – Assists in assessing model generalization, uncertainty analysis, and the role of catchment attributes.

 - Benchmarking & Regionalization – Facilitates large-scale model comparisons and parameter transfer to ungauged basins.

 - Hybrid Modeling – Enhances physics-based models with machine learning for bias correction and improved hydrologic simulations.

A notable study by Kratzert et al. (2019) demonstrated that LSTMs can outperform conceptual models in streamflow prediction. As part of this lab, we will explore how to programmatically download and load the data into R.

# What's in the data?

Each record in the CAMELS dataset represents a unique river basin, identified by an outlet USGS NWIS `gauge_id`. The dataset contains a mix of continuous and categorical variables, including meteorological, catchment, and streamflow summaries. 

The data we are going to download are the basin-level summaries. For example, if we looked at row 1 of the data (Gage: 01013500), all of the attribute values are areal averages over the drainage basin shown below, while the flow metrics are associated with the outlet gage (in red). We can use `dataRetrieval::findNLDI()` to fetch the basin polygon and upstream flowlines:

```{r}
#| message: false
#| eval: false

# Use the `findNLDI` function to get the basin and flowlines for the first gauge
basin <- dataRetrieval::findNLDI(nwis = "01013500", 
                                 # Navigate the "upper tributaries" of the basin
                                 nav = "UT", 
                                 # Return the basin and flowlines
                                 find = c("basin", "flowlines"))

# Plot the basin, flowlines, and gauge ...
ggplot() + 
  geom_sf(data = basin$basin, fill = "lightblue") + 
  geom_sf(data = basin$UT_flowlines, color = "blue") + 
  geom_sf(data = basin$origin, color = "red") + 
  theme_minimal()
```

## Lab Goals

This lab's worked example grounds the **minimum** tidymodels pipeline — split, recipe, workflow, fit, evaluate. Q3 and Q4 then ask you to extend that foundation with everything taught in the lectures.

By the end of the lab you will have:

 - Programmatically downloaded CAMELS data and joined basin-level attribute files into a modeling-ready frame.
 - Specified a `recipe`, dropped it into a `workflow`, and compared multiple models via `workflow_set` against cross-validated resamples *(Week 5-1)*.
 - Extracted **variable importance** and interpreted predictor ranks through the *geographic-proxy* lens *(Week 5-2)*.
 - Evaluated with **spatial cross-validation** and reported the honest gap between random and spatial CV *(Week 5-2)*.
 - **Tuned** hyperparameters with `tune_grid()` and a named grid, then finalized with `last_fit()` *(Week 6-1)*.
 - Produced a deployed Quarto page communicating the full pipeline.

The worked example below demonstrates (1) and a stripped-down (2). You'll apply the remaining capabilities — full workflow comparison, spatial CV, VIP, tuning, and `last_fit()` — in Q4 using the lecture material as reference.

# Lab Set Up

1. Make a new R project where you want this work to take place.
2. Create a Quarto document called `lab-05.qmd` in the project folder.
3. Create a `data` directory in your project folder.
4. Instantiate a git archive with `usethis::use_git()`
5. Connect it to a Github repository with `usethis::use_github()`

Let's start by loading the necessary libraries for this lab (you may need to install some!):

## Data Download

The CAMELS dataset is archived on [Zenodo](https://zenodo.org/records/15529996). Scroll down to the file listing — you will see a set of `.txt` files (climate, geology, soil, topography, vegetation, hydrology summaries per basin), along with a PDF describing every column. Each file has a direct-download link of the form `https://zenodo.org/records/15529996/files/<filename>?download=1`.

We will point all of our downloads at that root:

```{r}
root <- 'https://zenodo.org/records/15529996/files'
```

## Getting the documentation PDF

We can download the documentation PDF which provides descriptions for the various columns (many are not self-explanatory). We use `download.file` to pull it into our `data/` directory — binary files like PDFs need `mode = "wb"` on Windows.

```{r}
#| eval: false
download.file('https://zenodo.org/records/15529996/files/camels_attributes_v2.0.pdf?download=1',
              '../data/camels_attributes_v2.0.pdf',
              mode = 'wb')
```

## Getting Basin characteristics

Now we want to download the `.txt` files that store the actual data documented in the PDF.
Doing this file by file (like we did with the PDF) is possible, but lets look at a better/easier way...

a. Lets create a vector storing the data types/file names we want to download:

```{r}
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")
```

::: {.callout-warning collapse="true"}
### glue
The `glue` package provides an efficient way to interpolate and manipulate strings. It is particularly useful for dynamically constructing text, formatting outputs, and embedding R expressions within strings.

#### Key Features of glue:

 - **String Interpolation**: Embed R expressions inside strings using `{}`.
 - **Improved Readability**: Eliminates the need for cumbersome `paste()`, `paste0()` and `sprintf()` commands.
 - **Multi-line Strings**: Easily handle multi-line text formatting.
 - **Safe and Efficient**: Optimized for performance and readability.

#### Basic Usage
- To use `glue`, you need to load the package and then call the `glue()` function with the desired string template. You can embed R expressions within curly braces `{}` to interpolate values into the string.

```{r}
class <- "ESS 330"
year  <- 2024

glue("We are taking {class} together in {year}")
```

### Multiples

- You can also use glue to interpolate multiple values at once

```{r}
classes <- c("ESS 330", "ESS 523c")

glue("We are taking {classes} together in {year}")
```
:::

b. Using `glue`, we can construct the needed URLs and file names for the data we want to download. Zenodo direct-download links end with `?download=1`, so we append that to the remote URL but keep the local filename clean:

```{r}
# Where the files live online ...
remote_files  <- glue('{root}/camels_{types}.txt?download=1')
# where we want to download the data ...
local_files   <- glue('data/camels_{types}.txt')
```

c. Now we can download the data:
`walk2` comes from the `purrr` package and is used to apply a function to multiple arguments in parallel (much like `map2` works over paired lists). Here, we are asking `walk2` to pass the first element of `remote_files` and the first element of `local_files` to the `download.file` function to download the data, and setting `quiet = TRUE` to suppress output. The process is then iterated for the second element of each vector, and so on.

```{r}
#| eval: false
walk2(remote_files, local_files, download.file, quiet = TRUE)
```

d. Once downloaded, the data can be read it into R using `readr::read_delim()`, again instead of applying this to each file individually, we can use `map` to apply the function to each element of the `local_files` list.

```{r}
# Read and merge data
camels <- map(local_files, read_delim, show_col_types = FALSE) 
```

e. This gives us a list of data.frames, one for each file that we want to merge into a single table. So far in class we have focused on *_join functions to merge data based on a primary and foreign key relationship.

In this current list, we have >2 tables, but, have a shared column called `gauge_id` that we can use to merge the data. However, since we have more then a left and right hand table, we need a more robust tool. We will use the `powerjoin` package to merge the data into a single data frame. `powerjoin` is a flexible package for joining lists of `data.frames`. It provides a wide range of join types, including inner, left, right, full, semi, anti, and cross joins making it a versatile tool for data manipulation and analysis, and one that should feel familiar to users of `dplyr`.

In this case, we are join to merge every data.frame in the list (n = `r length(local_files)`) by the shared `gauge_id` column. Since we want to keep all data, we want a full join.

```{r}
camels <- power_full_join(camels ,by = 'gauge_id')
```

:::{.callout-note}
Alternatively, we could have read straight form the urls. Strongly consider the implications of this approach as the longevity and persistence of the data is not guaranteed.

```{r}
#| eval: false
# Read and merge data
camels <- map(remote_files, read_delim, show_col_types = FALSE) |> 
  power_full_join(by = 'gauge_id')
```
:::

# Question 1: Your Turn (15 points)

- [ ] Make sure all data and the PDF are downloaded into you data directory
- [ ] From the documentation PDF, report what `zero_q_freq` represents

## Exploratory Data Analysis

First, lets make a map of the sites. Use the `borders()` ggplot function to add state boundaries to the map and initially color the points by the mean flow (`q_mean`) at each site.

```{r}
#| warning: false
ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = q_mean)) +
  scale_color_gradient(low = "pink", high = "dodgerblue") +
  ggthemes::theme_map()
```

We can take a moment here to learn a few new things about `ggplot2`!

:::{.callout-note}
### Color scales
In `ggplot`, sometimes you want different things (like bars, dots, or lines) to have different colors. But how does R know which colors to use? That’s where color scales come in!

scales can be used to map data values to colors (`scale_color_*`) or fill aesthetics (`scale_fill_*`). There are two main types of color scales:

 - Discrete color scales – for things that are categories, like "apples," "bananas," and "cherries." Each gets its own separate color.
```{r}
#| eval: false
scale_color_manual(values = c("red", "yellow", "pink")) #lets you pick your own colors.
```

Or 

```{r}
#| eval: false
scale_color_brewer(palette = "Set1") #uses a built-in color set.
```

 - Continuous color scales – for numbers, like temperature (cold to hot) or height (short to tall). The color changes smoothly.

```{r}
#| eval: false
scale_color_gradient(low = "blue", high = "red") #makes small numbers blue and big numbers red.
```

:::

The zero_q_freq variable represents the frequency of days with Q = 0 mm/day, expressed as a percentage. It measures how often a basin's stream gauge records zero discharge. Values near 0% indicate perennial streams, while higher values indicate intermittent or ephemeral streams.

# Question 2: Your Turn (20 points)

- [ ] Make 2 maps of the sites, coloring the points by the `aridity` and `p_mean` columns
- [ ] Add clear labels, titles, and a color scale that makes sense for each parameter.
- [ ] Ensure these render as a single image with your choice of `facet_*`, `patchwork`, or `ggpubr`

```{r}
library(patchwork)

map_aridity <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = aridity)) +
  scale_color_gradient(low = "darkgreen", high = "red") +
  ggthemes::theme_map() +
  labs(title = "Aridity Across CAMELS Basins",
       color = "Aridity\n(PET/P)")

map_pmean <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = p_mean)) +
  scale_color_gradient(low = "lightyellow", high = "darkblue") +
  ggthemes::theme_map() +
  labs(title = "Mean Daily Precipitation Across CAMELS Basins",
       color = "Precip\n(mm/day)")

map_aridity / map_pmean
```


# Model Preparation

As an initial analysis, lets look at the relationship between aridity, rainfall and mean flow. First, lets make sure there is not significant correlation between these variables. Here, we make sure to drop NAs and only view the 3 columns of interest.

```{r}
camels |> 
  select(aridity, p_mean, q_mean) |> 
  drop_na() |> 
  cor()
```

As expected, there is a strong correlation between rainfall and mean flow, and an inverse correlation between aridity and rainfall. While both are high, we are going to see if we can build a model to predict mean flow using aridity and rainfall.

##  Visual EDA

a. Lets start by looking that the 3 dimensions (variables) of this data. We'll start with a XY plot of aridity and rainfall. We are going to use the `scale_color_viridis_c()` function to color the points by the `q_mean` column. This scale functions maps the color of the points to the values in the `q_mean` column along the viridis continuous (c) palette. Because a `scale_color_*` function is applied, it maps to the known color aesthetic in the plot.

```{r}
# Create a scatter plot of aridity vs rainfall
ggplot(camels, aes(x = aridity, y = p_mean)) +
  # Add points colored by mean flow
  geom_point(aes(color = q_mean)) +
  # Add a linear regression line
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  # Apply the viridis color scale
  scale_color_viridis_c() +
  # Add a title, axis labels, and theme (w/ legend on the bottom)
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

Ok! so it looks like there **is** a relationship between rainfall, aridity, and rainfall but it looks like an exponential decay function and is certainly not linear.

To test a transformation, we can log transform the x and y axes using the `scale_x_log10()` and `scale_y_log10()` functions:

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c() +
  # Apply log transformations to the x and y axes
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

Great! We can see a log-log relationship between aridity and rainfall provides a more linear relationship. This is a common relationship in hydrology and is often used to estimate rainfall in ungauged basins. However, once the data is transformed, the lack of spread in the streamflow data is quite evident with high mean flow values being compressed to the low end of aridity/high end of rainfall.

To address this, we can visualize how a log transform may benefit the `q_mean` data as well. Since the data is represented by color, rather than an axis, we can use the `trans` (transform) argument in the `scale_color_viridis_c()` function to log transform the color scale.

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  # Apply a log transformation to the color scale
  scale_color_viridis_c(trans = "log") +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom",
        # Expand the legend width ...
        legend.key.width = unit(2.5, "cm"),
        legend.key.height = unit(.5, "cm")) + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

Excellent! Treating these three right skewed variables as log transformed, we can see a more evenly spread relationship between aridity, rainfall, and mean flow. This is a good sign for building a model to predict mean flow using aridity and rainfall.

# Model Building

## Lets start by splitting the data

First, we set a seed for reproducibility, then transform the `q_mean` column to a log scale. Remember it is error prone to apply transformations to the outcome variable within a `recipe`. So, we'll do it a priori.

Once set, we can split the data into a training and testing set. We are going to use 80% of the data for training and 20% for testing. We are leaving stratification off in this minimal example, but as Week 5-1 covered, stratified splits often matter when the outcome has structure — you'll revisit this choice in Q4.

Additionally, we are going to create a 10-fold cross validation dataset to help us evaluate multi-model setups.

```{r}
set.seed(123)
# Bad form to perform simple transformations on the outcome variable within a 
# recipe. So, we'll do it here.
camels <- camels |> 
  mutate(logQmean = log(q_mean))

# Generate the split
camels_split <- initial_split(camels, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

camels_cv <- vfold_cv(camels_train, v = 10)
```

## Preprocessor: `recipe`

In Week 5-1 we saw how a `recipe` lets us declare a preprocessing pipeline once, estimate its parameters on the training data, and reuse it everywhere — folds, test set, new predictions.

We learned quite a lot about the data in the visual EDA. We know that the `q_mean`, `aridity`, and `p_mean` columns are right skewed and can be helped by log transformations. We also know that the relationship between `aridity` and `p_mean` is non-linear and can be helped by adding an interaction term. Let's build a small recipe that encodes those decisions (your Q4 recipe will be bigger):

```{r}
# Create a recipe to preprocess the data
rec <-  recipe(logQmean ~ aridity + p_mean, data = camels_train) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())
```

## Naive base `lm` approach:

Ok, to start, lets do what we are comfortable with ... fitting a linear model to the data. First, we use `prep` and `bake` on the training data to apply the `recipe.` Then, we fit a linear model to the data.

```{r}
# Prepare the data
baked_data <- prep(rec, camels_train) |> 
  bake(new_data = NULL)

# Interaction with lm
#  Base lm sets interaction terms with the * symbol
lm_base <- lm(logQmean ~ aridity * p_mean, data = baked_data)
summary(lm_base)

# Sanity Interaction term from recipe ... these should be equal!!
summary(lm(logQmean ~ aridity + p_mean + aridity_x_p_mean, data = baked_data))
```

## Where things get a little messy...

Ok so now we have our trained model `lm_base` and want to validate it on the test data.

> Remember a models ability to predict on new data is the most important part of the modeling process. It really doesnt matter how well it does on data it has already seen!

We have to be careful about how we do this with the base R approach:

### Wrong version 1: augment

The `broom` package provides a convenient way to extract model predictions and residuals. We can use the `augment` function to add predicted values to the test data. However, if we use `augment` directly on the test data, we will get incorrect results because the preprocessing steps defined in the `recipe` object have not been applied to the test data.

```{r}
#| error: true
nrow(camels_test)
nrow(camels_train)
broom::augment(lm_base, data = camels_test)
```

### Wrong version 2: `predict`

The `predict` function can be used to make predictions on new data. However, if we use `predict` directly on the test data, we will get incorrect results because the preprocessing steps defined in the `recipe` object have not been applied to the test data.

```{r}
#| error: true
camels_test$p2 = predict(lm_base, newdata = camels_test)

## Scales way off!
ggplot(camels_test, aes(x = p2, y = logQmean)) +
  geom_point() +
  # Linear fit line, no error bands
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  # 1:1 line
  geom_abline(color = "red", linewidth = 1) +
  labs(title = "Linear Model Using `predict()`",
       x = "Predicted Log Mean Flow",
       y = "Observed Log Mean Flow") +
  theme_linedraw()
```

### Correct version: `prep` -> `bake` -> `predict`

To correctly evaluate the model on the test data, we need to apply the same preprocessing steps to the test data that we applied to the training data. We can do this using the `prep` and `bake` functions with the `recipe` object. This ensures  the test data is transformed in the same way as the training data before making predictions.

```{r}
test_data <-  bake(prep(rec), new_data = camels_test)
test_data$lm_pred <- predict(lm_base, newdata = test_data)
```

## Model Evaluation: statistical and visual

Now that we have the predicted values, we can evaluate the model using the `metrics` function from the `yardstick` package. This function calculates common regression metrics such as RMSE, R-squared, and MAE between the observed and predicted values.

```{r}
metrics(test_data, truth = logQmean, estimate = lm_pred)

ggplot(test_data, aes(x = logQmean, y = lm_pred, colour = aridity)) +
  # Apply a gradient color scale
  scale_color_gradient2(low = "brown", mid = "orange", high = "darkgreen") +
  geom_point() +
  geom_abline(linetype = 2) +
  theme_linedraw() + 
  labs(title = "Linear Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

Ok so that was a bit burdensome, is really error prone (fragile), and is worthless if we wanted to test a different algorithm... lets look at a better approach!

## Using a `workflow` instead

`tidymodels` provides a framework for building and evaluating models using a consistent and modular workflow. The `workflows` package lets you bundle a **preprocessor** (recipe or formula) and a **model spec** into a single object, then fit them together. This makes it easier to experiment with different models, compare performance, and ensure reproducibility.

Here, we use `linear_reg()` to define a linear regression model, set the engine to `lm`, and the mode to `regression`. We then add our recipe to the workflow, fit the model to the training data, and extract the coefficients.

```{r}
# Define model
lm_model <- linear_reg() %>%
  # define the engine
  set_engine("lm") %>%
  # define the mode
  set_mode("regression")

# Instantiate a workflow ...
lm_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(lm_model) %>%
  # Fit the model to the training data
  fit(data = camels_train) 

# Extract the model coefficients from the workflow
summary(extract_fit_engine(lm_wf))$coefficients
```

Lets ensure we replicated the results from the lm_base model. How do they look to you?

```{r}
# From the base implementation
summary(lm_base)$coefficients
```

## Making Predictions

Now that `lm_wf` is a workflow, data is not embedded in the model, we can use augment with the new_data argument to make predictions on the test data.
 
```{r}
#
lm_data <- augment(lm_wf, new_data = camels_test)
dim(lm_data)
```

## Model Evaluation: statistical and visual

As with EDA, graphical and statistical evaluation of the model is key. Here, we use the `metrics` function to extract the default regression metrics (rmse, rsq, mae) between the observed and predicted mean streamflow values.

We then create a scatter plot of the observed vs predicted values, colored by aridity, to visualize the model performance.

```{r}
metrics(lm_data, truth = logQmean, estimate = .pred)

ggplot(lm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

## Switch it up!

The real power of this approach is that we can easily switch out the models/recipes and see how it performs. Here, we are going to instead use a random forest model to predict mean streamflow. We define a random forest model using the `rand_forest` function, set the engine to `ranger`, and the mode to `regression`. We then add the recipe, fit the model, and evaluate the skill.



```{r}
library(baguette)
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(rf_model) %>%
  # Fit the model
  fit(data = camels_train) 
```

## Predictions

- Make predictions on the test data using the `augment` function and the `new_data` argument.

```{r}
rf_data <- augment(rf_wf, new_data = camels_test)
dim(rf_data)
```

## Model Evaluation: statistical and visual

Evaluate the model using the `metrics` function and create a scatter plot of the observed vs predicted values, colored by aridity.

```{r}
metrics(rf_data, truth = logQmean, estimate = .pred)

ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

Awesome! We just set up a completely new model and were able to utilize all of the things we had done for the linear model. This is the power of the `tidymodels` framework!

That said, we still have some repetition — and, more importantly, we are eyeballing metrics on a single train/test split rather than comparing the two models head-to-head under a fair, resampled evaluation. `workflow_set()` lets us do both in a few lines.

## A `workflowset` approach

`workflow_set` is a powerful tool for comparing multiple models on the same data. It allows you to define a set of workflows, fit them to the same data, and evaluate their performance using a common metric. Here, we are going to create a `workflow_set` object with the linear regression and random forest models, fit them to the training data, and compare their performance using the `autoplot` and `rank_results` functions.

```{r}
wf <- workflow_set(list(rec), list(lm_model, rf_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv) 

autoplot(wf)

rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

Overall it seems the random forest model is outperforming the linear model. This is not surprising given the non-linear relationship between the predictors and the outcome :)

## Final Model

The workflow-set comparison above told us the random forest wins. Let's fit it once more on the full training set so we have a single model object to interrogate:

```{r}
final <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_model) |>
  fit(data = camels_train)
```

## Evaluation

A proper evaluation has three parts:

1. **Variable importance** — which predictors is the model actually using? Week 5-2 introduced `vip::vip()` paired with `extract_fit_parsnip()` to pull the underlying model object out of a fitted workflow.
2. **Test-set metrics** — apply the trained model to data it has never seen.
3. **A predicted-vs-observed plot** — the visual sanity check; the 1:1 line should run through the cloud.

```{r}
# Variable importance (Week 5-2 pattern)
final |>
  extract_fit_parsnip() |>
  vip::vip()

# Predictions on the test set
rf_data <- augment(final, new_data = camels_test)

# Regression metrics (rmse, rsq, mae by default)
metrics(rf_data, truth = logQmean, estimate = .pred)

ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  # 1:1 line — perfect prediction
  geom_abline(linetype = 2) +
  # Actual regression line for contrast
  geom_smooth(method = "lm", col = 'red', linetype = 2, se = FALSE) +
  theme_linedraw() +
  labs(title = "Random Forest Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

That's the **minimum** end-to-end loop. Q4 asks you to do the full version: swap in spatial CV, tune the hyperparameters, and close with `last_fit()` — all patterns the lectures cover in detail.

# Question 3: Your Turn! (40 points)

Extend the `workflow_set` above (which already contains the linear model and the random forest) by adding two more models from the Week 6-1 zoo:

 - [ ] Build an `xgboost` (engine) regression (mode) model using `boost_tree`
 - [ ] Build a neural network model using the `nnet` engine from the `baguette` package using the `bag_mlp` function
 - [ ] Add both to the workflow set and refit against the resamples
 - [ ] Compare all four (linear, random forest, xgboost, neural network) with `autoplot()` and `rank_results()`
 - [ ] Which of the four would you move forward with? Justify the pick in 2–3 sentences.
 
```{r}
# XGBoost model
xgb_model <- boost_tree() |>
  set_engine("xgboost") |>
  set_mode("regression")

# Neural network model
nn_model <- bag_mlp() |>
  set_engine("nnet") |>
  set_mode("regression")

# Rebuild workflowset with all four models
wf_all <- workflow_set(
  list(rec), 
  list(lm_model, rf_model, xgb_model, nn_model)
) |>
  workflow_map("fit_resamples", resamples = camels_cv)

# Compare all four
autoplot(wf_all)

rank_results(wf_all, rank_metric = "rsq", select_best = TRUE)
```
 
Based on the workflow set comparison, the bagged neural network performed best across the 10-fold cross validation, achieving the lowest RMSE and highest R² among all four models. The random forest was a close second, also outperforming the linear model and XGBoost. I would move forward with the bag_mlp model given its predictive performance, though the random forest is a strong alternative given its greater interpretability through variable importance scores.

# Bridge to Q4: what the lectures cover

The worked example above stops at a single random-forest fit with default hyperparameters. Q4 asks you to extend it with the rest of the unit. Before starting, re-skim the relevant lecture sections — you'll be lifting patterns directly from them:

| Q4 task | Lecture reference |
|---|---|
| EDA with `skimr`/`visdat`, normality checks, stratification | **Week 5-1** — *Exploratory Data Analysis*, *Data Normality*, *Stratification* |
| `recipe` design (impute, dummy, normalize, interact) | **Week 5-1** — *Feature engineering with `recipes`*, *Common normalization techniques* |
| Workflow sets across multiple models | **Week 5-1** — *Further simplification: `workflowsets`* |
| Variable importance & geographic-proxy interpretation | **Week 5-2** — Part 1 (*VIP*), *Which predictors carry geography?* |
| Spatial cross-validation (`spatial_block_cv`, `spatial_clustering_cv`, `spatial_leave_location_out_cv`) | **Week 5-2** — Part 3, *Applying this to your data*, *The same idea in regression* |
| Random-vs-spatial CV gap as a leakage diagnostic | **Week 5-2** — *The moment of truth*, *What a leaky model looks like*, *The 2×2 comparison* |
| Hyperparameter tuning, named grids, `finalize_workflow()`, `last_fit()` | **Week 6-1** — *Optimizing Models via Tuning*, *The final fit!* |

**If a line item in the rubric below doesn't look familiar, that's your cue to re-watch the relevant lecture slide — do not try to infer the API from this lab alone.**

# Question 4: Build your own (200 points)

Borrowing from the workflow presented above, build your own complete ML pipeline to predict mean streamflow using the CAMELS dataset.

This question asks you to apply the full toolkit from Week 5 and Week 6: EDA, recipes, workflow sets, **spatial cross-validation** (CAMELS is a spatial dataset — random CV will flatter your model), hyperparameter **tuning**, variable importance, and a clean final fit with `last_fit()`. Q4 is the bulk of the lab — budget time accordingly.

:::{.callout-tip}
**Benchmark**: a well-built model typically achieves **R² > 0.9 under random CV**. Under spatial CV, expect a 0.05–0.10 drop — that's honest generalization, not a failure. **Report both.** If spatial CV stays within ~0.05 of random CV, you have strong evidence your predictors carry transferable signal.
:::

## Data prep & EDA (30) — *Week 5-1*

 - [ ] Set a seed for reproducibility
 - [ ] Use `skimr::skim()` and/or `visdat::vis_dat()` on your working data. Describe — in 3–4 sentences — what this tells you about missingness, variable types, and distributions. Use this to justify decisions you make in the recipe below.
 - [ ] Visualize the distribution of `q_mean` (and your chosen predictors) — note anything skewed, bimodal, or bounded that will drive a `step_*` choice.
 - [ ] *Optional but encouraged*: run `ape::Moran.I()` on each numeric predictor (Week 5-2 slide *"Which predictors carry geography?"*) to see which features carry spatial structure. High Moran's I = candidate geographic proxy to revisit in the VIP step.
 - [ ] Create an initial split with 75% used for training and 25% for testing. Consider whether stratification (e.g., by `aridity` quartile or hydrologic class) is appropriate and **explain your choice**.
 - [ ] Extract your training and testing sets
 
```{r}
library(skimr)
library(visdat)

# Set seed for reproducibility
set.seed(123)

# Overview of the data
skim(camels)
vis_dat(camels)
```

```{r}
# Distribution of outcome variable
ggplot(camels, aes(x = q_mean)) +
  geom_histogram(bins = 50, fill = "steelblue") +
  labs(title = "Distribution of q_mean", x = "Mean Streamflow (mm/day)") +
  theme_linedraw()

# Distribution of key predictors
ggplot(camels, aes(x = aridity)) +
  geom_histogram(bins = 50, fill = "coral") +
  labs(title = "Distribution of Aridity") +
  theme_linedraw()

ggplot(camels, aes(x = p_mean))+
  geom_histogram(bins = 50, fill = "darkgreen") +
  labs(title = "Distribution of p_mean") +
  theme_linedraw()
```

 The vis_dat() output shows that the CAMELS dataset contains 671 observations, with the majority of variables being numeric and a small number of character variables representing categorical attributes like geologic class, dominant land cover, and precipitation timing. Missing values are sparse but present in several columns, primarily in soil and geology variables, which will be addressed using imputation in the recipe. Distribution plots reveal that q_mean, aridity, and p_mean are all right-skewed, justifying the log transformations applied in the worked example. These findings support using step_log() for skewed predictors and step_dummy() for categorical variables in the recipe.
 
```{r}
camels_split2 <- initial_split(camels, prop = 0.75, strata = aridity)
camels_train2 <- training(camels_split2)
camels_test2  <- testing(camels_split2)

# Check the sizes
nrow(camels_train2)
nrow(camels_test2)

```
 

I chose to stratify the split by aridity because its distribution is strongly right-skewed with a few extreme values above 3. Without stratification, those extreme arid basins could end up entirely in one split, making the training and test sets unrepresentative of the full range of conditions. Stratifying ensures both sets contain a proportional mix of wet and dry basins.

## Recipe (30) — *Week 5-1*

 - [ ] Define a formula you want to use to predict `logQmean`
 - [ ] In 5–7 sentences, justify your predictor choice. Consult the downloaded PDF; ground **each** recipe step (log, normalize, interact, naomit, impute, ...) in what the EDA above showed.
 - [ ] Build a recipe that handles those predictors. Your recipe should contain **at least three** `step_*` calls beyond the formula, and each should be there for a reason you can name.
 
```{r}
rec2 <- recipe(logQmean ~ aridity + p_mean + pet_mean + 
                 slope_mean + elev_mean + soil_porosity + 
                 soil_conductivity + max_water_content + 
                 sand_frac + frac_forest + frac_snow,
               data = camels_train2) |>
  step_log(aridity, p_mean, pet_mean) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_interact(terms = ~ aridity:p_mean) |>
  step_naomit(all_predictors(), all_outcomes())
```
 
 Predictors were selected to represent the major controls on mean streamflow across the water, energy, and landscape domains. Aridity, p_mean, and pet_mean capture the climatic water balance — the EDA showed all three are right-skewed, justifying step_log(). Elevation and slope reflect topographic controls on runoff generation, while soil porosity, conductivity, and max water content 
represent the subsurface's capacity to store and transmit water. Sand fraction and forest cover influence infiltration and evapotranspiration respectively.  step_impute_median() handles the sparse missing values identified in the vis_dat() output. step_normalize() is applied because we plan to compare models (including neural networks and SVMs) that are sensitive to predictor 
scale. Finally, the aridity-precipitation interaction term captures the non-linear relationship clearly visible in the worked example EDA.

## Define 3 models (30) — *Week 6-1*

 - [ ] Define a random forest model using the `rand_forest` function (engine `ranger`, mode `regression`)
 - [ ] Define two other models of your choice. The Week 6-1 zoo (`linear_reg`, `boost_tree`, `svm_rbf`, `svm_poly`, `mlp`, `bag_mlp`, `decision_tree`) is a good menu — the comparison table in that deck summarizes strengths and weaknesses.
 - [ ] For each model, briefly note (one sentence) why it's a reasonable candidate for this regression problem — what kind of structure does it capture?

```{r}
# Model 1: Random Forest
rf_model2 <- rand_forest() |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")

# Model 2: XGBoost
xgb_model2 <- boost_tree() |>
  set_engine("xgboost") |>
  set_mode("regression")

# Model 3: Neural Network
nn_model2 <- bag_mlp() |>
  set_engine("nnet") |>
  set_mode("regression")
```

The random forest model is a strong candidate because it handles non-linear relationships and interactions between predictors naturally, which is important given the complex relationships between climate, soil, and streamflow. XGBoost is appropriate because it builds trees sequentially to correct errors, making it effective at capturing subtle patterns in tabular hydrologic data. The 
bagged neural network is included because it can learn complex non-linear mappings between predictors and streamflow that neither tree-based method may fully capture.

## Workflow set — compare with *honest* CV (35) — *Week 5-2*

With your preprocessing steps and models defined, build a `workflow_set` and compare the models across resamples. Because CAMELS basins are spatially distributed, a random `vfold_cv` can over-state performance if your model leans on geographic proximity — the Week 5-2 deck walks through this carefully, including when the gap is real and when it isn't.

 - [ ] Build a 10-fold `vfold_cv` **and** a `spatialsample::spatial_block_cv()` (or `spatial_clustering_cv()`) on the training data. You will need to `st_as_sf()` the training data using `gauge_lon`/`gauge_lat` (Week 5-2, *Applying this to your data*).
 - [ ] Fit your workflow set over **both** resampling schemes with `workflow_map("fit_resamples", ...)`
 - [ ] Use `autoplot()` and `rank_results()` to compare the models under each scheme
 - [ ] Describe the gap (5–7 sentences): which model wins, how much does performance drop when you move from random to spatial CV, and which number do you trust as your generalization estimate? What does the spread of per-fold metrics tell you about *where* your model struggles? Recall from lecture that a **small** gap is a signal that your predictors carry real transferable signal — not a failed diagnostic.

```{r}
library(sf)
library(spatialsample)

# Random 10-fold CV 
camels_cv2 <- vfold_cv(camels_train2, v = 10)

# Spatial CV 
camels_sf <- sf::st_as_sf(camels_train2, 
                      coords = c("gauge_lon", "gauge_lat"),
                      crs = 4326)

# Spatial block CV
camels_spatial_cv <- spatial_block_cv(camels_sf, v = 10)
```


```{r}
# Build the workflow set
wf_set2 <- workflow_set(
  list(rec2),
  list(rf_model2, xgb_model2, nn_model2)
)

# Fit against random CV
wf_random <- wf_set2 |>
  workflow_map("fit_resamples", 
               resamples = camels_cv2,
               verbose = TRUE)

# Fit against spatial CV
wf_spatial <- wf_set2 |>
  workflow_map("fit_resamples", 
               resamples = camels_spatial_cv,
               verbose = TRUE)
```

```{r}
# Compare random CV results
autoplot(wf_random) + 
  labs(title = "Model Comparison: Random CV")

rank_results(wf_random, rank_metric = "rsq", select_best = TRUE)

# Compare spatial CV results
autoplot(wf_spatial) + 
  labs(title = "Model Comparison: Spatial CV")

rank_results(wf_spatial, rank_metric = "rsq", select_best = TRUE)
```


Under spatial cross-validation, the bagged neural network (bag_mlp) performed best with an R² of 0.900 and RMSE of 0.340, followed by random forest (R² = 0.881) and XGBoost (R² = 0.858). This ranking differed slightly from random CV where bag_mlp also led with R² = 0.937. All models showed a modest drop of 0.04-0.06 R² units when moving to spatial CV, suggesting our predictors carry genuine transferable hydrologic signal. XGBoost showed the largest drop (0.056), suggesting it may rely more on spatial proximity than the other models. I trust the spatial CV estimate as the honest generalization metric since CAMELS 
basins span diverse climatic regions across the US.

```{r}
rank_results(wf_spatial, rank_metric = "rsq", select_best = TRUE)
```

```{r}
rank_results(wf_spatial, rank_metric = "rsq", select_best = TRUE) |>
  select(wflow_id, model, .metric, mean, rank)
```

## Tune the best model (25) — *Week 6-1*

 - [ ] Tag **at least two** hyperparameters with `tune()` in your chosen model specification
 - [ ] Build a named grid (e.g. `grid_space_filling()`, `grid_regular()`) — do not just pass an integer to `tune_grid()`. Justify your grid choice in one sentence.
 - [ ] Run `tune_grid()` against your **spatial** resamples with `control_grid(save_pred = TRUE)`
 - [ ] Show the tuning landscape with `autoplot()` and identify the best combination with `show_best()` / `select_best()`
 - [ ] In 2–3 sentences, describe which hyperparameters mattered most and what you learned about the model's sensitivity
 
```{r}
# Random forest with tunable hyperparameters
rf_tune <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("regression")
```

```{r}
# Build tuning workflow
rf_tune_wf <- workflow() |>
  add_recipe(rec2) |>
  add_model(rf_tune)

# Build a space filling grid - better coverage than random grid
rf_grid <- grid_space_filling(
  mtry(range = c(2, 11)),
  min_n(range = c(2, 20)),
  size = 20
)

# Run tuning against spatial CV
rf_tune_results <- tune_grid(
  rf_tune_wf,
  resamples = camels_spatial_cv,
  grid = rf_grid,
  control = control_grid(save_pred = TRUE)
)
```

```{r}
autoplot(rf_tune_results)
show_best(rf_tune_results, metric = "rsq")
```
 
The tuning landscape reveals that mtry had a stronger influence on model performance than min_n. Very low mtry values produced higher RMSE and lower R², while values around 4-6 performed best. The min_n parameter showed relatively flat performance 
across its range, suggesting the model is not highly sensitive to this setting. The best combination was mtry = 4 and min_n = 9, achieving a spatial CV R² of 0.882.

## Variable importance (25) — *Week 5-2*

 - [ ] Produce a `vip::vip()` plot for your tuned model (pattern: `extract_fit_parsnip() |> vip()`)
 - [ ] Interpret the top predictors (4–6 sentences). Cross-reference against the predictor Moran's I you computed in EDA — are any of the top VIP features *also* highly spatially autocorrelated? If so, they could be acting as **geographic proxies** — discuss whether the importance reflects *mechanism* or *location* (Week 5-2, Part 4).
 - [ ] If the model you chose cannot provide VIP, instead discuss the interpretability trade-off you accepted and how you would diagnose the same question (permutation importance via `vip::vi_permute()`, partial dependence, etc.).
 
```{r}
# Select the best hyperparameters
best_params <- select_best(rf_tune_results, metric = "rsq")

# Finalize the workflow with best params
final_wf <- finalize_workflow(rf_tune_wf, best_params)

# Fit on full training data
final_fit <- fit(final_wf, data = camels_train2)

# Variable importance plot
final_fit |>
  extract_fit_parsnip() |>
  vip::vip(num_features = 15) +
  labs(title = "Variable Importance: Tuned Random Forest") +
  theme_linedraw()
```

The VIP plot reveals that p_mean and aridity are by far the most important predictors of mean streamflow, with importance scores roughly double that of all other variables combined. This makes strong physical sense, the water balance between precipitation inputs and evaporative demand is the primary control on how much water exits a basin as streamflow. However, both p_mean and aridity are also highly spatially structured across the US (wet in the east/northwest, dry in the southwest), meaning they may partly act as geographic proxies, encoding location rather than pure mechanism. The interaction term aridity_x_p_mean ranks fourth, confirming the non-linear relationship between 
these two variables identified in the EDA. Forest fraction ranks third, reflecting its role in evapotranspiration, forested basins tend to have lower streamflow relative to precipitation. Soil variables ranked lowest, suggesting that at this large sample scale, climate dominates over subsurface properties in controlling mean flow.

## Final fit & evaluate (25) — *Week 6-1*

Use the canonical tidymodels closing move (Week 6-1, *The final fit!*):

 - [ ] `finalize_workflow()` your tuned workflow with the best hyperparameters
 - [ ] Run `last_fit()` against your original `initial_split()` object — this trains on the full training set and evaluates on the held-out test set in one call
 - [ ] Use `collect_metrics()` and `collect_predictions()` to report final test-set performance
 - [ ] Create a plot of observed vs predicted `logQmean` on the **test set** with clear title, axis labels, and a compelling color scale (e.g., colored by aridity or region). Include a 1:1 reference line.
 - [ ] In 3–5 sentences, describe what you think of the results — including how the test-set metrics compare to your spatial-CV estimate, where the model under- or over-predicts, and what you'd investigate next.

```{r}
# Finalize and run last_fit against the original split
final_last_fit <- last_fit(final_wf, camels_split2)

# Collect metrics and predictions
collect_metrics(final_last_fit)

# Plot observed vs predicted
collect_predictions(final_last_fit) |>
  ggplot(aes(x = logQmean, y = .pred)) +
  geom_point(aes(color = camels_test2$aridity)) +
  scale_color_viridis_c() +
  geom_abline(linetype = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  theme_linedraw() +
  labs(title = "Final Model: Observed vs Predicted Log Mean Flow",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

The final tuned random forest model achieved an R² of 0.908 and RMSE of 0.343 on the held-out test set, performing very close to the spatial CV estimate of R² = 0.882 — a gap of only 0.026, suggesting the model generalizes well to unseen basins. The observed vs predicted plot shows strong agreement for mid-to-high flow basins (dark purple, low aridity), where points cluster tightly around the 1:1 line. The model struggles most with very dry, high-aridity basins (teal and yellow points in the bottom left), where it tends to overpredict streamflow, likely because intermittent and ephemeral streams are harder to characterize with the predictors available. To improve performance in arid basins, future work could include zero_q_freq as a predictor or explore 
separate models for perennial vs intermittent stream regimes.

# Rubric

 - [ ] **Q1**:  Download data (15 pts)
 - [ ] **Q2**:  Make 2 maps (20 pts)
 - [ ] **Q3**:  Add xgboost + neural network to the workflow set (40 pts)
 - [ ] **Q4a**: Data prep & EDA (30 pts)
 - [ ] **Q4b**: Recipe with justification (30 pts)
 - [ ] **Q4c**: Define 3 models (30 pts)
 - [ ] **Q4d**: Workflow set + spatial CV comparison (35 pts)
 - [ ] **Q4e**: Tuning (named grid, ≥ 2 hyperparameters, autoplot) (25 pts)
 - [ ] **Q4f**: VIP interpretation with geographic-proxy lens (25 pts)
 - [ ] **Q4g**: `last_fit()` evaluation & plot (25 pts)
 - [ ] **Well-structured, compelling Qmd deployed as web page** (25 pts)

**Total: 300 points** (Q4 = 200 pts / 67%)

# Submission:

Once you have completed the above steps, you can submit your lab for grading. The lab should be submitted as a URL to your deployed repo page. The repo should contain the `lab-05.qmd` file with all the code, data, and output.

Good luck!
