---
title: 'Lab 6: Timeseries Data'
format:
   html:
    toc: true
    author:
  - name: Taia Storrs
    email: taia.storrs@colostate.edu
---

# Introduction

In this lab you will download streamflow data from the Cache la Poudre River (USGS site 06752260) and analyze it using the time series methods introduced in lecture. You will then use `modeltime` to predict future streamflow using both the time structure of the series and exogenous climate predictors.

This mirrors the Lees Ferry forecasting example from lecture — same `dataRetrieval` pipeline, same `modeltime` workflow, but now applied to a river you can see from campus. Like Lees Ferry, the Poudre is an operational water-supply system: Fort Collins, Loveland, and Greeley all depend on it, and the flows you forecast here directly inform how water managers plan deliveries.

```{r}
#| message: false
#| eval: false

install.packages("exactextractr")
install.packages("tidymodels")
install.packages("tsibble")
install.packages("feasts")
install.packages("modeltime")
install.packages("timetk")
install.packages("tseries")
```

```{r}
options(download.file.method = "curl")
remotes::install_github("mikejohnson51/climateR")
```


```{r}
#| message: false
library(tidyverse)
library(plotly)

library(dataRetrieval)
library(climateR)
library(terra)
library(exactextractr)

library(tidymodels)
library(tsibble)
library(feasts)
library(modeltime)
library(timetk)
library(tseries)   # for stationarity tests
```

## Getting the Data

### Streamflow

Download daily streamflow for the Poudre at Mouth (USGS 06752260) and aggregate to monthly means. This is the same `readNWISdv()` + `renameNWISColumns()` pipeline from Lab 1 — the only addition is `yearmonth()` from `tsibble` to create a monthly time index.

```{r}
#| message: false
#| warning: false
poudre_flow <- readNWISdv(
  siteNumber = "06752260",
  parameterCd = "00060",
  startDate   = "2013-01-01",
  endDate     = "2023-12-31"
) |>
  renameNWISColumns() |>
  mutate(Date = yearmonth(Date)) |>
  group_by(Date) |>
  summarise(Flow = mean(Flow))
```

### Historic Climate (GridMET)

We will use `climateR::getTerraClim()` to download monthly climate data for the Poudre basin (max temperature, precipitation, solar radiation). The `exactextractr` package spatially averages each raster layer over the basin polygon.

```{r}
#| message: false
#| warning: false
basin <- findNLDI(nwis = "06752260", find = "basin")

sdate <- as.Date("2013-01-01")
edate <- as.Date("2023-12-31")

gm <- getTerraClim(
  AOI       = basin$basin,
  var       = c("tmax", "ppt", "srad"),
  startDate = sdate,
  endDate   = edate
) |>
  unlist() |>
  rast() |>
  exact_extract(basin$basin, "mean", progress = FALSE)

historic <- mutate(gm, id = "gridmet") |>
  pivot_longer(cols = -id) |>
  mutate(name = sub("^mean\\.", "", name)) |>
  tidyr::extract(name, into = c("var", "index"), "(.*)_([^_]+$)") |>
  mutate(index = as.integer(index)) |>
  mutate(Date = yearmonth(seq.Date(sdate, edate, by = "month")[as.numeric(index)])) |>
  pivot_wider(id_cols = Date, names_from = var, values_from = value) |>
  right_join(poudre_flow, by = "Date")

# Validate: should have 132 rows (11 years × 12 months). If fewer, climate data
# is missing for some months and those flow months were silently dropped.
stopifnot("Climate join dropped flow months — check GridMET download" = nrow(historic) == 132)
```

### Future Climate (MACA)

Any time exogenous predictors are used for forecasting, **future values of those predictors must also be available**. We use MACA — a statistically downscaled climate dataset developed by the same group as GridMET — to get projected temperature, precipitation, and solar radiation through 2033.

Two differences from GridMET: temperature is in Kelvin (subtract 273.15 for °C) and variable names differ slightly.

```{r}
#| message: false
#| warning: false
sdate <- as.Date("2024-01-01")
edate <- as.Date("2033-12-31")

maca <- getMACA(
  AOI      = basin$basin,
  var      = c("tasmax", "pr", "rsds"),
  timeRes  = "month",
  startDate = sdate,
  endDate   = edate
) |>
  unlist() |>
  rast() |>
  exact_extract(basin$basin, "mean", progress = FALSE)

future <- mutate(maca, id = "maca") |>
  pivot_longer(cols = -id) |>
  mutate(name = sub("^mean\\.", "", name)) |>
  tidyr::extract(name, into = c("var", "index"), "(.*)_([^_]+$)") |>
  mutate(index = as.integer(index)) |>
  mutate(Date = yearmonth(seq.Date(sdate, edate, by = "month")[as.numeric(index)])) |>
  pivot_wider(id_cols = Date, names_from = var, values_from = value)

# Rename the 3 climate columns to match GridMET names (Date column stays first)
stopifnot("Unexpected MACA column count" = ncol(future) == 4)
names(future)[2:4] <- c("ppt", "srad", "tmax")  # order matches getMACA var request
future <- mutate(future, tmax = tmax - 273.15)   # Kelvin → °C
```

------------------------------------------------------------------------

# Your Turn

## 1. Convert to tsibble

Convert `historic` to a `tsibble` object using `as_tsibble()`. A `tsibble` is a tidy time series data frame — it retains the date index and integrates with `feasts` functions for decomposition and visualization. Think of it as a tibble that knows it is ordered in time.

```{r}
#| eval: false
#| echo: false
pf <- as_tsibble(historic)
```

## 2. Plot the Time Series

Use `ggplot` to plot monthly Flow over time. Use `plot_ly()` from `plotly` for an interactive version — this lets you zoom into specific years or hover over values.

```{r}
#| eval: false
#| echo: false
g1 <- ggplot(pf, aes(x = as.Date(Date), y = Flow)) +
  geom_line(color = "steelblue") +
  labs(title  = "Monthly Streamflow — Cache la Poudre at Mouth",
       y = "Flow (cfs)", x = NULL) +
  theme_minimal()

plot_ly(pf, x = ~as.Date(Date), y = ~Flow, type = "scatter", mode = "lines",
        line = list(color = "steelblue")) |>
  layout(title = "Monthly Streamflow — Cache la Poudre at Mouth",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Flow (cfs)"))
```

## 3. Seasonal Diagnostics

Before fitting any model, inspect the seasonal structure using two complementary `feasts` plots.

**Subseries plot** — one panel per month, each line is one year, the blue line is the monthly mean across all years:

```{r}
#| eval: false
#| echo: false
gg_subseries(pf, y = Flow) +
  labs(title = "Monthly Flow Patterns — Poudre at Mouth",
       y = "Flow (cfs)", x = "Year") +
  theme_minimal()
```

**Seasonal plot** — each line is one year plotted over the calendar cycle. Parallel lines = stable seasonal shape; diverging lines = changing pattern:

```{r}
#| eval: false
#| echo: false
gg_season(pf, y = Flow) +
  labs(title = "Seasonal Shape by Year",
       y = "Flow (cfs)", x = "Month") +
  theme_minimal()
```

Describe what you see. How are "seasons" defined in these plots? Is the seasonal shape consistent across years, or does it change? What physical process drives the May–June peak?

The seasonal shape is consistent across years with low winter baseflow, a rapid rise in the spring with a sharp peak in May/June, and a decline through summer. However, the magnitude varies dramatically year to year. 2014 and 2015 have much bigger peaks (around 2000 cfs) while the drier years are just hitting 500 cfs. The May–June peak is driven by snowmelt. In the subseries plot, the blue lines confirm May and June are by far the highest-flow months on average, and the swings in the May/June panels show how unpredictable peak flows are year to year

## 4. Stationarity and Autocorrelation

Two diagnostics that directly inform ARIMA model selection:

**Lag plot** — plots Flow against its own lagged values. Strong linear patterns = high autocorrelation (ARIMA has something to work with). A diffuse cloud = no memory. Use `log1p(Flow)` — on the raw scale, peak months compress all other observations into the corner and hide the structure.

```{r}
#| eval: false
#| echo: false
pf |>
  mutate(log_flow = log1p(Flow)) |>
  gg_lag(y = log_flow, geom = "point") +
  labs(title = "Lag Structure of Monthly Poudre Flow (log scale)",
       x = "lag(log Flow, n)", y = "log Flow") +
  theme_minimal()
```

**Stationarity tests** — ARIMA requires a stationary series (constant mean and variance). Test formally before fitting:

```{r}
#| eval: false
#| echo: false
# Test on log_flow — that's what ARIMA will actually model
log_flow_vec <- log1p(pf$Flow)

# ADF: H0 = non-stationary; p < 0.05 rejects non-stationarity → series IS stationary
adf.test(log_flow_vec)

# KPSS: H0 = stationary; p < 0.05 rejects stationarity → series is NOT stationary
kpss.test(log_flow_vec)
```

What do the tests tell you? If the series is non-stationary, ARIMA will need `d ≥ 1` (differencing). Does the lag plot support the same conclusion?

Lag 1 shows a strong positive diagonal. Knowing this month's flow allows for strong predictive power for next month. However, the structure gets messier as lags increase. The middle lags are hard to predict because we're bridging across the seasonal transition. Structure comes back around lag 9. The color clustering confirms the strong seasonal signal. ADF having p<0.01 means  reject non-stationarity, which tells us that the series is stationary. KPSS having p=0.04 means reject stationarity, which tells us that the series is not stationary. The tests disagreeing tells us that the ADF is picking up the strong mean-reversion, while KPSS is flagging the seasonal variance as non-stationary behavior. auto.arima() will likely use d = 0 but may add seasonal differencing (D = 1) to handle the seasonal component.

## 5. Decomposition

Use `model(STL(...))` to decompose the time series into trend, seasonality, and remainder. Choose a trend window size you feel is appropriate and justify your choice. Extract components with `components()` and visualize with `autoplot()`.

```{r}
#| eval: false
#| echo: false
decomp <- pf |>
  model(STL(Flow ~ trend(window = 12) + season(window = "periodic"))) |>
  components()

autoplot(decomp)
```

Describe what you see:

-   What does the **trend** tell you about long-term changes in Poudre flow? Is there evidence of a decline consistent with climate change or upstream demand?
-   What drives the **seasonal** component? (Hint: snowmelt, irrigation withdrawals, reservoir releases)
-   Are there any spikes in the **remainder** that correspond to anomalous years you can identify?

The trend rises from ~100 cfs in 2013 up to a peak around ~600 cfs in 2015-2016, then declines steadily through 2019-2020 bottoming out near ~50 cfs, before a modest recovery toward 2023. This is consistent with the western drought cycle, the late 2010s were severe drought years across Colorado, reducing overall basin wetness and snowpack. The seasonal component is a near-identical repeating spike every year. The peak each year represents snowmelt-driven runoff, while the negative values in winter reflect base flow falling below the annual mean. Irrigation withdrawals and reservoir operations also suppress late-summer flows below what snowmelt alone would predict. The large positive spike in late 2013 is almost certainly the September 2013 flood. The large negative spikes around 2015-2016 suggest the trend component slightly overestimated flow during the transition out of the wet period. After 2018 the remainder is much flatter, meaning the models explain the data well in drier years. 

------------------------------------------------------------------------

## Modeltime Prediction

### Data Preparation

Convert the historic tsibble and future tibble to plain tibbles with a `Date` column for `modeltime`. All predictor columns must be present in both the historic and future datasets — this is what enables the 10-year climate-driven forecast.

Streamflow is strictly non-negative but right-skewed — snowmelt peaks dwarf base-flow months. To prevent models from predicting negative flows, add a **log-transformed response** with `log1p(Flow)` (`log(Flow + 1)`, safe when flow is near zero). Fit all models on `log_flow`; back-transform predictions with `expm1()` (= `exp(x) - 1`) to return to cfs units.

```{r}
#| eval: false
#| echo: false
pf_tbl <- pf |>
  as_tibble() |>
  mutate(date     = as.Date(Date),
         log_flow = log1p(Flow),
         Date     = NULL)

future_tbl <- future |>
  as_tibble() |>
  mutate(date = as.Date(Date), Date = NULL)
```

### Train / Test Split

Hold out the last 24 months (2022–2023) as the test set. This mirrors the `assess = "60 months"` logic from the CO₂ and Lees Ferry examples — the test set is always the most recent data, never a random sample.

```{r}
#| eval: false
#| echo: false
# Note: time_series_split() is deterministic — it always cuts at the same point
# regardless of the seed. set.seed() is kept here as a tidymodels convention
# (some resampling functions are random), but removing it gives identical results.
set.seed(10291991)
splits <- time_series_split(pf_tbl, assess = "24 months", cumulative = TRUE)

training <- training(splits)
testing  <- testing(splits)
```


### Model Definition

Choose at least three models — one ARIMA and one Prophet are required. Store model specifications in a list:

```{r}
#| eval: false
#| echo: false
mods <- list(
  # Classical seasonal ARIMA — relies on stationarity (check your tests above)
  arima_reg() |> set_engine("auto_arima"),

  # ARIMA + XGBoost residual boosting — picks up nonlinear climate relationships
  arima_boost() |> set_engine("auto_arima_xgboost"),

  # Prophet — handles trend breaks, missing data, and multiple seasonalities
  prophet_reg() |> set_engine("prophet"),

  # Prophet + XGBoost — as above, but boosts on residuals
  prophet_boost() |> set_engine("prophet_xgboost"),

  # Exponential smoothing — strong baseline for multiplicative seasonal data
  exp_smoothing() |> set_engine("ets"),

  # MARS — nonparametric spline model, useful for nonlinear climate response
  mars(mode = "regression") |> set_engine("earth")
)
```

### Model Fitting

Fit all models to the training data on the log scale. Exclude the raw `Flow` column so models only see `log_flow` as the response — each engine handles `date` and climate predictors internally.

```{r}
#| eval: false
#| echo: false
models <- map(mods, ~ fit(.x, log_flow ~ ., data = select(training, -Flow)))

(models_tbl <- as_modeltime_table(models))
```

### Calibration and Accuracy

Calibrate on the test set (2022–2023) to compute out-of-sample accuracy metrics. Pass `select(testing, -Flow)` so calibration uses the log-scale response:

```{r}
#| eval: false
#| echo: false
calibration_table <- modeltime_calibrate(models_tbl, select(testing, -Flow), quiet = FALSE)

modeltime_accuracy(calibration_table) |>
  arrange(mae) |>
  select(.model_desc, mae, mape, mase, rmse, rsq)
```

Describe what you see. Use these benchmarks to interpret the numbers:

-   **MASE \< 1** — the model beats a naïve seasonal forecast (last year's same month)
-   **MAPE** — intuitive % error, but inflated in low-flow months when Flow ≈ 0
-   **RMSE** — penalizes large errors heavily; compare to mean test-period flow for context
-   Do boosted models (ARIMA-boost, Prophet-boost) outperform their base versions? If so, the nonlinear climate signal is worth capturing.

EARTH (MARS) wins with the lowest MAE (0.448) and highest R² (0.849), and the nonparametric spline model handled the nonlinear climate-flow relationships best. ETS comes in second (MAE 0.511). ARIMA-boost third, Prophet models in the middle, and Prophet-boost last. 

All 6 models have MASE < 1. Boosted models did not consistently outperform base models; ARIMA-boost beat plain ARIMA, but Prophet-boost was the worst performer. This suggests that the climate signal is somewhat noisy. MAPE of 13–18% is reasonable for monthly streamflow  and R² of 0.73–0.85 means models explain 73–85% of variance in log-flow during the test period. 

### Test Set Forecast

Forecast over the held-out test period and compare predictions to actuals. Back-transform with `expm1()` to return predictions to cfs. Two details matter: (1) `expm1(x)` is the exact inverse of `log1p(x)` and always returns values ≥ 0; (2) confidence interval lower bounds on the log scale can be slightly negative — floor them at 0 with `pmax(.conf_lo, 0)` *before* back-transforming so the ribbon never dips below zero cfs:

```{r}
#| eval: false
#| echo: false
test_forecast <- calibration_table |>
  modeltime_forecast(new_data = select(testing, -Flow), actual_data = pf_tbl) |>
  mutate(.value   = expm1(.value),
         .conf_lo = expm1(pmax(.conf_lo, 0)),
         .conf_hi = expm1(.conf_hi))

test_forecast |>
  filter(.key == "actual") |>
  ggplot(aes(x = .index, y = .value)) +
  geom_line(color = "grey40") +
  geom_line(data = filter(test_forecast, .key == "prediction"),
            aes(color = .model_desc)) +
  geom_ribbon(data = filter(test_forecast, .key == "prediction"),
              aes(ymin = .conf_lo, ymax = .conf_hi, fill = .model_desc),
              alpha = 0.1) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(title  = "Test Set Forecast vs Actuals — Poudre at Mouth",
       x = NULL, y = "Flow (cfs)", color = "Model", fill = "Model") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Do predictions track actuals during the 2022–2023 test window? Are confidence intervals capturing observed values? Note any discontinuities at the 2022 boundary between training and test.

All 6 models are clustered tightly together in the test period, which makes sense given the similarity in their accuracy metrics. All models also correctly captured the seasonal timing, however, all models underestimated the 2023 peak. The 2022 peak was predicted reasonably well as models got close to the ~1200 cfs actual. The ARIMA-boost ribbon is extremely wide, but most other models have tight, overlapping ribbons. The wide intervals do capture the 2023 peak, but at the cost of being uninformatively broad. There's a visible discontinuity at the 2022 boundary where predictions start, and no models predicted negative flow. 

### Refit to Full Dataset

Refit all models to the complete 2013–2023 record before forecasting forward. Refitting allows auto-tuned models (`auto.arima`, `ets`) to re-optimize their parameters with more data:

```{r}
#| eval: false
#| echo: false
refit_tbl <- modeltime_refit(calibration_table, data = select(pf_tbl, -Flow))

modeltime_accuracy(refit_tbl) |>
  arrange(mae) |>
  select(.model_desc, mae, mape, mase, rmse, rsq)
```

Do the accuracy metrics change after refitting? Why or why not?

The metrics didn't change because refit accuracy still reflects test-set performance, not in-sample metrics. The UPDATE: tag confirms models were re-trained on the full dataset, it kept the same structure. Stable rankings gives us confidence that EARTH and ETS are robustly the best performers, not just lucky on a particular train/test cut. 

### Forecast into the Future (2024–2033)

Use the refitted models with `future_tbl` as `new_data` to generate a 10-year climate-conditioned forecast. Back-transform with `expm1()`:

```{r}
#| eval: false
#| echo: false
future_forecast <- refit_tbl |>
  modeltime_forecast(new_data = future_tbl, actual_data = pf_tbl) |>
  mutate(.value   = expm1(.value),
         .conf_lo = expm1(pmax(.conf_lo, 0)),
         .conf_hi = expm1(.conf_hi))

future_forecast |>
  filter(.key == "actual") |>
  ggplot(aes(x = .index, y = .value)) +
  geom_line(color = "grey40") +
  geom_line(data = filter(future_forecast, .key == "prediction"),
            aes(color = .model_desc)) +
  geom_ribbon(data = filter(future_forecast, .key == "prediction"),
              aes(ymin = .conf_lo, ymax = .conf_hi, fill = .model_desc),
              alpha = 0.1) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  labs(title  = "10-Year Forecast — Poudre at Mouth (2024–2033)",
       x = NULL, y = "Flow (cfs)", color = "Model", fill = "Model") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

------------------------------------------------------------------------

## Wrap-up

Examine your 10-year forecast and reflect on the following:

1.  **Model agreement**: Do ARIMA and Prophet project similar trajectories? If they diverge, what does that tell you about their different assumptions (statistical stationarity vs. flexible trend)?

ARIMA and Prophet project similar flat/modest trajectories, reflecting their shared assumption of a stable historical pattern. EARTH diverges dramatically because it learned aggressive nonlinear relationships between climate variables and flow that extrapolate unrealistically far into the future.

2.  **Forecast quality**: Which model would you recommend to CWCB for water-supply planning? Justify with your accuracy metrics.

ETS or plain ARIMA for operational planning, both had strong test-set MASE (<0.52), produce conservative forecasts consistent with historical ranges, and have interpretable uncertainty bounds. EARTH's diverging forecast makes it unsuitable despite winning on test metrics.

3.  **Negative predictions**: If any model predicts negative flow, explain why this happens and what you would do to fix it in a production context (hint: log-transformation of Flow before modeling).

No negatives appeared here because log-transformation kept all predictions on a positive scale. expm1() always returns values ≥ 0.

4.  **Regressor value**: The MACA climate projections assume a specific emissions scenario. How does uncertainty in those projections propagate into your streamflow forecast?

MACA uses a specific emissions scenario. If actual emissions differ, the projected temperature and precipitation inputs will be wrong, and errors in those predictors compound directly into streamflow forecast errors. Models like EARTH that rely heavily on climate regressors inherit this uncertainty most severely.

5.  **Improvement path**: What information not in `tmax`, `ppt`, `srad` would most improve this forecast? (Consider: snowpack, upstream storage, irrigation demand.)

Snowpack would be the single biggest improvement, since it directly measures how much water will enter the river during melt season. Upstream reservoir storage levels and irrigation demand schedules would also help capture human-managed flow components.

At this stage you have a working end-to-end time series forecasting pipeline — the same structural approach used by CWCB, USBR, and state water agencies for operational planning. The processes you learned in `tidymodels` apply directly here, and you can extend this framework with cross-validation (`time_series_cv()`), feature engineering (lag variables, rolling means), and hyperparameter tuning.

------------------------------------------------------------------------

## Going Further (Optional): Add an Upstream Regressor

In lecture we showed that adding Glenwood Springs flow as an exogenous regressor improved Lees Ferry forecasts because upstream conditions carry physical information the model can't infer from time structure alone. Try the same approach here.

Download data from a headwater gage on the Poudre (e.g., North Fork Cache la Poudre, USGS 06746095) and add it as a predictor:

```{r}
#| eval: false
upstream <- readNWISdv(
  siteNumber  = "06746095",
  parameterCd = "00060",
  startDate   = "2013-01-01",
  endDate     = "2023-12-31"
) |>
  renameNWISColumns() |>
  # floor_date returns the 1st of each month as a Date — same type as pf_tbl$date
  mutate(date = floor_date(Date, "month")) |>
  group_by(date) |>
  summarise(upstream_flow = mean(Flow, na.rm = TRUE), .groups = "drop")

pf_reg <- pf_tbl |>
  left_join(upstream, by = "date") |>
  drop_na(upstream_flow)

splits_reg <- time_series_split(pf_reg, assess = "24 months", cumulative = TRUE)

# Use log_flow ~ . consistent with the main pipeline; exclude raw Flow
models_reg <- map(mods, ~ fit(.x, log_flow ~ .,
                               data = select(training(splits_reg), -Flow)))
```

Does adding upstream flow improve MASE on the test set? If yes, the physical signal is worth including. If not, the climate data alone may be sufficient for monthly aggregation at this timescale.
