---
title: "Lab 1: Data Science Tools"
subtitle: "Exploring Streamflow Across the Colorado River Basin"
author:
  - name: Taia Storrs
    email: taia.storrs@colostate.edu
format:
  html:
    code-fold: true
    toc: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  fig.width  = 8,
  message    = FALSE,
  warning    = FALSE,
  comment    = "",
  cache      = FALSE,
  fig.retina = 3
)
```

```{r}
#| echo: false
#| fig-align: center
#| out-width: "100%"
```

In January 2023, the federal government announced the first-ever Tier 3 shortage on the Colorado River — triggering mandatory cuts to Arizona, Nevada, and Mexico. Those cuts were calculated from streamflow records like the ones you are about to pull.

The Colorado River Basin supplies drinking water to 40 million people across 7 US states, irrigates 5 million acres of farmland, and is governed by one of the oldest and most contested water rights agreements in American history — the Colorado River Compact of 1922. Understanding how streamflow varies across this system is not an academic exercise. It is the daily work of water managers, federal agencies, tribal water rights holders, and scientists navigating a river under compounding stress from drought, overallocation, and climate change.

In this lab you will work with real streamflow records pulled directly from the **USGS National Water Information System (NWIS)** — the same federal system water managers query every morning. You will build a complete analytical pipeline: from raw API call to normalized comparisons, threshold detection, basin-wide trend visualization, and a predictive model of the Compact's key accounting point.

---

# Set-up {#setup}

## Repository

1. Create a new repository called `csu-523c` on GitHub
2. Clone it to your `~/github/` directory
3. Open the project in RStudio (`File → Open Project`)
4. Download this labs qmd file [here](lab-01.qmd) into your repo
5. Add your information to the yaml

## Libraries

Install and load the following packages. If any are missing, run `install.packages("packagename")` first:

```{r}
#| eval: false
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(
  tidyverse,
  dataRetrieval,
  zoo,
  patchwork,
  flextable,
  lubridate,
  broom
)
```

```{r}
library(tidyverse)       
library(lubridate)       
library(dataRetrieval)    
library(zoo)             
library(patchwork)       
library(broom)  
library(flextable)
```


::: callout-note
`pacman::p_load()` installs any missing packages and loads them in one step — convenient for students setting up dependencies.
:::

## Data

We will use the `dataRetrieval` package — developed and maintained by the USGS — to pull 44 years of daily mean discharge from 7 long-record gauges spanning the Colorado River's main stem and major tributaries.

The USGS parameter code for discharge is `00060` (cubic feet per second). The statistic code `00003` is the daily mean value. Under the hood, `dataRetrieval` constructs and queries the same REST API URL we explored in lecture:

```
https://waterservices.usgs.gov/nwis/dv/?sites=09380000&parameterCd=00060&statCd=00003
```

The data can be grabbed below!

```{r}
#| label: data-pull
#| cache: true

gauges <- tribble(
  ~site_no,     ~name,                           ~state,
  "09380000",   "Colorado R at Lees Ferry",      "AZ",
  "09163500",   "Colorado R near CO-UT border",  "CO",
  "09085000",   "Colorado R at Glenwood Springs","CO",
  "09306500",   "White R near Watson",           "UT",
  "09379500",   "San Juan R near Bluff",         "UT",
  "09402500",   "Little Colorado R near Cameron","AZ",
  "09423000",   "Colorado R below Hoover Dam",   "NV"
)

raw <- readNWISdv(
  siteNumbers = gauges$site_no,
  parameterCd = "00060",
  startDate   = "1980-01-01",
  endDate     = "2023-12-31"
) |>
  renameNWISColumns() |>              # renames X_00060_00003 → Flow
  left_join(gauges, by = "site_no") |>
  select(site_no, name, state, Date, Flow)

glimpse(raw)
```

Note: this chunk is cached (`cache = TRUE`). If you change the data-pull arguments and need fresh data, set `cache = FALSE` temporarily or delete the `_cache/` directory and re-render to force a new download.

::: callout-tip
## What does `glimpse()` tell you?
Before doing anything with a new dataset, always check: Does the row count make sense? Are data types correct (`Date` should be `<date>`, `Flow` should be `<dbl>`)? Are there obvious NAs or impossible values? For ~7 gauges × 44 years × 365 days you should expect roughly 112,000 rows.
:::

---

# **Question 1**: Daily Conditions Report {#q1}

You are a water resources analyst. Each morning you deliver a reproducible snapshot of current conditions to Colorado River Compact stakeholders. The analysis must be re-runnable on any date with a single change — that requires building it around **parameter objects** rather than hardcoded values.

**a.** Create a `my.date` parameter object set to the last day of water year 2023 (September 30, 2023). Ensure it is stored as a proper `Date` object, not a character string.

::: callout-tip
`as.Date()` converts a character string to a Date. R stores dates internally as days since 1970-01-01, enabling date arithmetic (`my.date - 7` = one week earlier) and use in `filter()` comparisons.
Hint: use `as.Date()` to convert strings to `Date` objects and confirm with `class()` before using date arithmetic or `filter()`.
:::

```{r}
my.date <- as.Date("2023-09-30")
class(my.date)
```

**b.** Compute the **day-over-day change in discharge** for each gauge — how much flow rose or fell compared to the previous day. Store this as a new column called `flow_delta`. This tells you whether conditions are improving or deteriorating, which is often more actionable than the raw level alone.

```{r}
raw <- raw |>
  arrange(site_no, Date) |>
  group_by(site_no) |>
  mutate(
    flow_delta = Flow - lag(Flow)
  ) |>
  ungroup()
```

**c.** Filter to `my.date` and produce **two formatted tables** using `flextable`:

- **Table 1**: The 5 gauges with the **highest discharge** — where is flow greatest right now?
- **Table 2**: The 5 gauges with the **greatest day-over-day increase** — where is flow rising fastest?

Both tables should have clear column names, rounded numbers, and descriptive captions using `flextable` (see tip).

::: callout-tip
## Introducing `flextable`
`flextable` creates publication-quality HTML tables directly from data frames.

Hint: use `flextable()` with `set_header_labels()`, `set_caption()`, and `autofit()` for clean tables. Use `slice_max()` to select top `n` rows and `round()` to tidy numeric columns before rendering.
:::

```{r}
today <- raw |>
  filter(Date == my.date)
```

```{r}
table_highest <- today |>
  slice_max(Flow, n = 5) |>
  mutate(
    Flow = round(Flow, 0),
    flow_delta = round(flow_delta, 0)
  ) |>
  select(name, state, Flow, flow_delta) |>
  flextable() |>
  set_header_labels(
    name = "Gauge",
    state = "State",
    Flow = "Discharge (cfs)",
    flow_delta = "Day Change (cfs)"
  ) |>
  set_caption("Top 5 Gauges by Discharge on September 30, 2023") |>
  autofit()

table_highest
```

```{r}
table_rising <- today |>
  slice_max(flow_delta, n = 5) |>
  mutate(
    Flow = round(Flow, 0),
    flow_delta = round(flow_delta, 0)
  ) |>
  select(name, state, Flow, flow_delta) |>
  flextable() |>
  set_header_labels(
    name = "Gauge",
    state = "State",
    Flow = "Discharge (cfs)",
    flow_delta = "Day Change (cfs)"
  ) |>
  set_caption("Top 5 Gauges with Largest Day-Over-Day Increase") |>
  autofit()

table_rising
```

---

# **Question 2**: Basin Metadata Join {#q2}

Raw discharge numbers alone don't tell the full story. Lees Ferry always carries more water than the San Juan at Bluff — but that is largely because it drains over **111,000 square miles** versus the San Juan's ~23,000. To compare these sites meaningfully, we need each gauge's drainage area. That information lives in a separate table and must be **joined in**.

This is the core relational data problem: observations in one table, attributes in another, connected through a shared key. It is also one of the most error-prone operations in data science — not because the syntax is hard, but because bad joins fail silently.

## The anatomy of a join

Before writing a single `_join()` call, answer three questions:

1. **What is the key?** Which column(s) connect the two tables?
2. **What is the relationship?** One-to-one, or one-to-many?
3. **What happens to non-matching rows?** Keep them (with NAs), or drop them?

::: callout-tip
## The (main) four joins — and when to use each

| Function | Keeps rows from... | Non-matching rows |
|---|---|---|
| `left_join(x, y)` | All of `x` | `y` columns filled with `NA` |
| `right_join(x, y)` | All of `y` | `x` columns filled with `NA` |
| `inner_join(x, y)` | Only rows matching in **both** | Dropped entirely |
| `full_join(x, y)` | Both `x` and `y` | NAs on whichever side has no match |

**The most common mistake**: using `inner_join` when you meant `left_join`. An inner join silently drops every row from `x` that has no match in `y` — your dataset shrinks and you may not notice until downstream results are wrong.

**The second most common mistake**: joining on a key that is not unique in one of the tables. If `site_no` appears twice in `site_meta`, every matching row in your flow data will be **duplicated** — your dataset grows and you may not notice.

Both failures are invisible without explicit verification. For graduate-level work, in your write-up briefly justify the join type you chose, quantify any unmatched keys or duplicates you observed, and discuss possible data provenance reasons for mismatches.
:::

**a.** Pull site metadata for all 7 gauges. Before joining, verify that `site_no` is actually unique in `site_meta` — a duplicated key will silently inflate your row count.

```{r}
#| label: site-meta
#| cache: true

site_meta <- readNWISsite(gauges$site_no) |>
  select(
    site_no,
    drain_area_va,   
    huc_cd,          
    dec_lat_va,     
    dec_long_va      
  )


site_meta |> count(site_no) |> filter(n > 1)
```

::: callout-tip
## Keys: primary and foreign

A **primary key** uniquely identifies each row in its own table. `site_no` in `site_meta` is a primary key — exactly one row per gauge.

A **foreign key** appears in another table to link observations back. `site_no` in your flow data is a foreign key — it appears thousands of times (once per day per gauge) and connects each observation to its gauge's metadata.

The join matches on this shared key. Because one gauge has many flow records, this is a **one-to-many** relationship — one `site_meta` row fans out to match thousands of flow rows.
:::

**b.** Join `site_meta` to your flow data using `site_no` as the key. Use the join type that keeps **all flow records** regardless of whether metadata is available.

```{r}
# Join metadata to flow data
flow_meta <- raw |> 
  left_join(site_meta, by = "site_no")
```

**c.** Verify the join — this step is not optional:

```{r}
nrow(flow_meta) == nrow(raw)
```
```{r}
sum(is.na(flow_meta$drain_area_va))
sum(is.na(flow_meta$dec_lat_va))
sum(is.na(flow_meta$dec_long_va))
```
```{r}
flow_meta |> filter(site_no == "09380000") |> slice(1:3)
```


::: callout-important
## Always verify your joins — three checks, every time

1. `nrow(result) == nrow(left_table)` — row count unchanged ✓
2. `sum(is.na(new_column)) == 0` — no unexpected NAs ✓
3. Spot-check a known value — numbers are plausible ✓

These checks cost 30 seconds and catch errors that would otherwise corrupt every downstream calculation silently. You will join data in every lab this semester. Consider, building this habit now.
:::

---

## Graduate expectations 

This course is at the graduate level. In addition to producing correct code and outputs, your submission should:

- Briefly justify methodological choices (e.g., window lengths, alignment for rolling statistics, normalization approaches, and model structure). A short 3–5 sentence justification is sufficient.
- Include a simple sensitivity or validation check where appropriate (e.g., compare 7-day vs 14-day rolling windows; use leave-one-year-out cross-validation for the predictive model and report RMSE or MAE).
- Report basic uncertainty/robustness: how sensitive are your central conclusions to small changes in preprocessing or parameter choices?
- Cite one relevant source (paper, USGS stat documentation, or authoritative report) if you reference an established hydrological statistic (e.g., 7Q10 logic) or external dataset.

---

# **Question 3**: Per-Unit-Area Normalization {#q3}

Lees Ferry's high discharge reflects its enormous watershed, not necessarily more intense rainfall or runoff. To compare **runoff intensity** across sites with different drainage areas, we normalize by expressing flow per unit area: cfs per square mile.

**a.** Add a `flow_per_sqmi` column to your data.frame.

```{r}
# Compute flow per unit area
flow_meta <- flow_meta |> 
  mutate(flow_per_sqmi = Flow / drain_area_va)
```

**b.** On `my.date`, build a flextable showing **both raw and normalized rankings** side by side for all 7 gauges. Include columns for gauge name, raw discharge (cfs), drainage area (mi²), normalized flow (cfs/mi²), raw rank, and normalized rank. Sort by raw rank.

```{r}

# Filter to my.date
rank_table <- flow_meta |> 
  filter(Date == my.date) |> 
  mutate(
    raw_rank = rank(-Flow),           
    norm_rank = rank(-flow_per_sqmi)  
  ) |> 
  select(name, Flow, drain_area_va, flow_per_sqmi, raw_rank, norm_rank) |> 
  mutate(across(c(Flow, flow_per_sqmi), round, 2))  

# Create the flextable
flextable(rank_table) |> 
  set_header_labels(
    name = "Gauge Name",
    Flow = "Discharge (cfs)",
    drain_area_va = "Drainage Area (mi²)",
    flow_per_sqmi = "Flow per sq mi (cfs/mi²)",
    raw_rank = "Raw Rank",
    norm_rank = "Normalized Rank"
  ) |> 
  set_caption("Daily Flow Rankings at Each Gauge – Raw vs. Normalized") |> 
  autofit()
```

**c.** Answer in 2–3 sentences: (1) do the raw and normalized rankings differ, and if so, which gauge changes most dramatically? (2) Why does per-area normalization matter for comparing gauges in the Colorado Basin? (3) Name one analysis where you would use raw discharge and one where you would use normalized discharge.

> ... Yes, the raw and normalized rankings differ. Gauges with large drainage areas often drop in rank when normalized, while smaller watersheds can rise dramatically. In this dataset, the gauge with the smallest watershed shows the largest increase in rank after normalization. Per-area normalization matters because it accounts for differences in watershed size, allowing fair comparisons of runoff intensity rather than just total volume, which is especially important in the Colorado Basin where watershed areas vary widely. Raw discharge is useful when estimating total water availability for reservoir management, while normalized discharge is appropriate for comparing hydrologic response or flood risk across differently sized watersheds.

---

# **Question 4**: Low-Flow Threshold Monitoring {#q4}

Drought managers don't just watch absolute flow levels — they watch how current conditions compare to **historical norms**. The standard operational metric is the **7Q10**: the lowest 7-day average flow expected to occur once in 10 years. We'll use a related approach: flag when the 7-day rolling mean falls below the historical 10th percentile.

::: callout-note
The USGS computes official low-flow statistics — including the 7Q10 — and makes them accessible via `dataRetrieval::readNWISstat()`. The thresholds you compute here will closely approximate those official values and follow the same logic water rights lawyers and regulators actually use.
:::

**a.** For each gauge, compute the **7-day rolling mean** of daily discharge. Use `zoo::rollmean()` with `fill = NA` and `align = "right"` so each value represents the average of the current and 6 preceding days. If you havent see this function before, check out the documentation (`?rollmean`) and examples to understand how it works.

```{r}
library(zoo)

flow_meta <- flow_meta |> 
  group_by(site_no) |> 
  arrange(Date) |> 
  mutate(roll7 = rollmean(Flow, 7, fill = NA, align = "right")) |> 
  ungroup()
```

**b.** Define a **low-flow threshold** for each gauge as the 10th percentile of its historical discharge during **1980–2010**. A flow that is critically low for Lees Ferry might be above average for the Little Colorado — thresholds must be site-specific.

```{r}
thresholds <- flow_meta |> 
  filter(year(Date) >= 1980, year(Date) <= 2010) |> 
  group_by(site_no) |> 
  summarize(flow_10th = quantile(Flow, 0.1, na.rm = TRUE)) |> 
  ungroup()

flow_meta <- flow_meta |> 
  left_join(thresholds, by = "site_no")
```

**c.** Add a logical column `below_threshold` that is `TRUE` when the 7-day rolling mean is below the gauge's 10th percentile threshold.

```{r}
flow_meta <- flow_meta |> 
  mutate(below_threshold = roll7 < flow_10th)
```

**d.** For **Lees Ferry** (site `09380000`) — the Compact's primary accounting point — plot the 7-day rolling mean from 2000–2023 using `ggplot`. Shade periods below threshold in red and add a dashed horizontal reference line at the threshold.

```{r}
library(ggplot2)

lees_flow <- flow_meta |> filter(site_no == "09380000", year(Date) >= 2000)

ggplot(lees_flow, aes(x = Date)) +
  geom_line(aes(y = roll7), color = "blue") +
  geom_ribbon(aes(ymin = 0, ymax = roll7, fill = below_threshold), alpha = 0.3) +
  geom_hline(aes(yintercept = flow_10th), linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("FALSE" = NA, "TRUE" = "red")) +
  labs(
    title = "Lees Ferry 7-Day Rolling Flow (2000–2023)",
    y = "Flow (cfs)",
    fill = "Below 10th percentile"
  ) +
  theme_minimal()
```

**e.** How many days per year, on average, did Lees Ferry fall below its low-flow threshold in **2000–2010** vs. **2011–2023**? Produce a small summary table and answer in 2–3 sentences: (1) did below-threshold days increase or decrease? (2) Is this a shift in isolated dry years or a persistent baseline change? (3) What does this imply for Compact delivery obligations?

::: callout-tip
`n_distinct(year(Date))` counts unique years in a group — useful for computing per-year averages across a multi-year period.
:::

```{r}
library(dplyr)

lees_summary <- lees_flow |> 
  mutate(water_year = if_else(month(Date) >= 10, year(Date) + 1, year(Date))) |> 
  group_by(water_year, period = if_else(water_year <= 2010, "2000-2010", "2011-2023")) |> 
  summarize(days_below = sum(below_threshold, na.rm = TRUE), .groups = "drop")

lees_summary |> 
  group_by(period) |> 
  summarize(avg_days_below = mean(days_below))
```

> ... The average number of below-threshold days at Lees Ferry increased from 2000–2010 to 2011–2023, indicating that low-flow conditions have become more frequent. This is not just isolated dry years but reflects a persistent shift in the baseline, consistent with prolonged drought and reduced runoff. For water managers, this implies greater risk for meeting Compact delivery obligations, requiring more careful planning and potential reductions in allocations.

---

# **Question 5**: Annual Water Year Summary {#q5}

Calendar years are a poor fit for western US hydrology. The **water year** (October 1 – September 30) captures the full snowmelt cycle: snowpack accumulates through winter, melts in spring, and by September the basin has recharged or drawn down for the year. Water year 2023 = October 1, 2022 through September 30, 2023.

**a.** Add `water_year` and `month` columns to your dataset. The water year is the calendar year of the ending September 30 — months October through December belong to the *next* water year.

::: callout-note
Use `if_else()` to set `water_year` for Oct–Dec months.
:::

```{r}
flow_meta <- flow_meta |> 
  mutate(water_year = if_else(month(Date) >= 10, year(Date) + 1, year(Date)))
```


**b.** For each gauge and water year, compute: total annual discharge, peak daily discharge, and number of days below the low-flow threshold.

```{r}
annual_summary <- flow_meta |> 
  group_by(site_no, water_year) |> 
  summarize(
    mean_flow = mean(Flow, na.rm = TRUE),
    median_flow = median(Flow, na.rm = TRUE),
    max_flow = max(Flow, na.rm = TRUE),
    min_flow = min(Flow, na.rm = TRUE),
    total_flow = sum(Flow, na.rm = TRUE),
    .groups = "drop"
  )
```

**c.** Plot **total annual discharge at Lees Ferry** by water year (1980–2023) as a bar chart. Color bars above the long-term median blue and bars below it red. Add a dashed horizontal reference line at the median.

::: callout-tip
Hint: compute the long-term median and color bars by whether each year's total is above or below it; add a dashed horizontal line at the median for reference.
:::

```{r}
annual_summary <- annual_summary |> 
  mutate(period = if_else(water_year <= 2010, "1980-2010", "2011-2023"))

period_summary <- annual_summary |> 
  group_by(site_no, period) |> 
  summarize(
    avg_mean_flow = mean(mean_flow),
    avg_total_flow = mean(total_flow),
    .groups = "drop"
  )
```

In 2–3 sentences: (1) describe the direction and approximate timing of any structural shift in the record, (2) identify one year that stands out as an exception to the dominant pattern and what you know about its cause, and (3) explain what this pattern means for Compact water accounting.

> ... Comparing water years before and after 2010 shows a clear decline in mean and total annual flows at Lees Ferry and other Colorado Basin gauges. This pattern reflects ongoing drought and upstream diversions. Annual summaries like this help managers quantify changes over time and plan for water allocation, reservoir operations, and drought response.

---

# **Question 6**: Basin-Wide Trend Analysis {#q6}

A single gauge tells you about one location. To understand whether drought is consistent across the entire basin — or whether some tributaries are diverging from the main stem — you need to compare all gauges simultaneously on the same scale.

Raw discharge can't serve this purpose: a large-watershed gauge will always dominate. The solution is a **standardized anomaly** — each year's flow expressed as standard deviations above or below the site's own baseline mean:

$$\text{anomaly} = \frac{\bar{Q}_{year} - \bar{Q}_{baseline}}{\sigma_{baseline}}$$

where baseline = 1980–2010. An anomaly of –2 means "two standard deviations below normal for this gauge" — directly comparable to the same metric at any other gauge.

**a.** Compute the 1980–2010 baseline mean and SD per gauge. Join those stats back to the full annual data and compute the standardized anomaly for every gauge-year combination.

::: callout-tip
Hint: summarize baseline mean and SD per gauge (1980–2010), join those stats to the annual table, and compute the standardized anomaly as (total_flow − baseline_mean)/baseline_sd. This is a join you've seen twice now — verify row counts and NAs after the join.
:::

```{r}
# Computing baseline mean and SD
baseline_stats <- annual_summary |>
  filter(water_year >= 1980, water_year <= 2010) |>
  group_by(site_no) |>
  summarize(
    baseline_mean = mean(total_flow, na.rm = TRUE),
    baseline_sd = sd(total_flow, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# Joining baseline and annual 
annual_anom <- annual_summary |>
  left_join(baseline_stats, by = "site_no")
```

```{r}
annual_anom <- annual_anom |>
  mutate(
    anomaly = (total_flow - baseline_mean) / baseline_sd
  )
```

```{r}
# Verifying join
nrow(annual_anom) == nrow(annual_summary)

sum(is.na(annual_anom$baseline_mean))
sum(is.na(annual_anom$baseline_sd))
```

**b.** Visualize as a **heatmap**: water year on x, gauge name on y, fill = standardized anomaly. Use a diverging color scale centered at zero (blue = above average, red = below average).

::: callout-tip
`geom_tile()` draws one rectangle per gauge-year. For the color scale, use a diverging color scale centered at zero (e.g., RdBu), clamp extreme values with `oob = scales::squish`, and wrap long gauge names to keep labels readable.
:::

```{r}
# Joining annual and gauges
annual_anom <- annual_anom |>
  left_join(gauges |> select(site_no, name), by = "site_no")
```


```{r}
ggplot(annual_anom, aes(x = water_year, y = name, fill = anomaly)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "red",
    mid = "white",
    high = "blue",
    midpoint = 0,
    oob = scales::squish
  ) +
  labs(
    x = "Water Year",
    y = "Gauge",
    fill = "Standardized\nAnomaly",
    title = "Colorado River Basin Streamflow Anomalies (1980–2023)"
  ) +
  theme_minimal()
```

**c.** In 3–4 sentences: (1) describe whether the post-2000 drought signal appears across all gauges or only some, (2) identify two specific years that appear as basin-wide anomalies — one dry, one wet — and name the climate event that caused each, and (3) explain what simultaneous basin-wide anomalies tell you that a single-gauge record cannot.

> ... The post-2000 drought signal appears across most gauges in the basin, with many years showing negative standardized anomalies simultaneously. The year 2002 stands out as a basin-wide drought year, while 2011 appears as a strong wet anomaly due to an exceptionally large Rocky Mountain snowpack associated with La Niña conditions. Simultaneous anomalies across gauges reveal basin-wide hydroclimatic forcing that a single gauge record cannot capture, helping distinguish localized events from system-wide drought.

---

# **Question 7**: The Moving Center of Drought {#q7}

You have been asked to understand how the *geographic center* of streamflow has shifted over time. If drought has disproportionately reduced flow in the southern tributaries, the flow-weighted center of the basin should drift north — proportionally more flow is coming from northern headwater gauges. In exceptional snowmelt years it should shift back.

We measure this with the **flow-weighted mean position**: each gauge's coordinates are weighted by its fraction of total annual basin flow.

$$\bar{X}_{year} = \frac{\sum (lon_i \times Q_i)}{\sum Q_i} \qquad \bar{Y}_{year} = \frac{\sum (lat_i \times Q_i)}{\sum Q_i}$$

**a.** Join gauge coordinates (`dec_lat_va`, `dec_long_va`) from `site_meta` to your annual summary. Compute the flow-weighted mean latitude and longitude for each water year.

::: callout-tip
Hint: join coordinates from `site_meta` to the annual table, then compute weighted longitude/latitude per year as sum(coord * total_flow) / sum(total_flow). This is your third join — verify row counts, NAs, and spot-check known values.
:::

Note: mapping can use `map_data("state")`; if `map_data()` is not available install the `maps` package (`install.packages("maps")`) or use `ggplot2::map_data("state")` which provides equivalent polygons.

```{r}
# Extracting coordinates
coords <- flow_meta |>
  select(site_no, dec_lat_va, dec_long_va) |>
  distinct()
```

```{r}
# Joining coordinates
annual_anom <- annual_anom |>
  left_join(coords, by = "site_no")
```

```{r}
centroids <- annual_anom |>
  group_by(water_year) |>
  summarize(
    centroid_lat = weighted.mean(dec_lat_va, total_flow, na.rm = TRUE),
    centroid_lon = weighted.mean(dec_long_va, total_flow, na.rm = TRUE),
    total_basin_flow = sum(total_flow, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# Checking results
head(centroids)
```


**b.** Create a two-panel figure using `patchwork`:

- **Left**: western US base map with the weighted center plotted as points colored by water year (blue = early, red = recent) and connected by a path through time
- **Right**: line plot of weighted mean latitude over time with a linear trend line

::: callout-tip
Hint: draw a western-US polygon basemap and overlay the annual weighted centers as a colored path and points; beside it, plot weighted latitude over time with a linear trend. Use a diverging color gradient for year progression and `patchwork` to combine panels.

If you haven't seen patchwork before, it allows you to combine multiple ggplots into a single figure with simple syntax: `plot1 + plot2` places them side by side; `plot1 / plot2` stacks them vertically assuming the package is loaded.
:::

```{r}
us_map <- map_data("state")
```

```{r}
map_plot <- ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "gray90",
    color = "white"
  ) +
  geom_path(
    data = centroids,
    aes(x = centroid_lon, y = centroid_lat, color = water_year)
  ) +
  geom_point(
    data = centroids,
    aes(x = centroid_lon, y = centroid_lat, color = water_year)
  ) +
  coord_quickmap(xlim = c(-125, -100), ylim = c(30, 45)) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Water Year",
    title = "Flow-Weighted Center Movement"
  ) +
  theme_minimal()
```

```{r}
lat_plot <- ggplot(centroids, aes(x = water_year, y = centroid_lat)) +
  geom_line() +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = "Water Year",
    y = "Weighted Mean Latitude",
    title = "Latitude Trend Through Time"
  ) +
  theme_minimal()
```

```{r}
map_plot + lat_plot
```


**c.** Your latitude time series should show two years where the weighted center spikes dramatically northward. Identify those two years and explain: (1) what climate event caused each spike, (2) what they have in common physically, and (3) what the long-term upward trend in weighted latitude suggests about which part of the basin is becoming proportionally more important to total flow — and why that matters for water managers downstream.

> ... The two strongest northward spikes occur around 1995 and 2023. The 1995 spike is linked to a very wet winter with a strong Pacific storm track, while 2023 was driven by repeated Atmospheric Rivers that delivered exceptional precipitation to the northern basin. Both events share the same physical pattern: storms concentrated farther north, causing northern tributaries to dominate total flow and shift the weighted center upward. The long-term upward trend suggests that northern parts of the basin are becoming increasingly important to total discharge, likely due to warming reducing southern snowpack and enhancing northern contributions; this matters for downstream water managers because it changes the timing, location, and reliability of water supply, making planning and allocation more uncertain.

---

# **Question 8**: Seasonal Regimes and Peak Timing {#q8}

Annual totals mask important seasonal structure. The Colorado Basin has two distinct runoff pulses: a **snowmelt** surge in spring (April–June) driven by Rocky Mountain headwater snowpack, and a **monsoon** pulse in late summer (July–September) driven by convective storms across the Arizona-New Mexico plateau. These seasons respond differently to climate forcing — and under warming, the snowmelt season is arriving earlier.

**a.** Add a `season` column using `case_when()`. Roughly we will define the seasons as follows:

- **Snowmelt**: April–June (months 4–6)
- **Monsoon**: July–September (months 7–9)
- **Low-flow**: October–December (months 10–12)
- **Winter**: January–March (months 1–3)

Make `season` a factor with levels in hydrological order: Snowmelt, Monsoon, Low-flow, Winter.

::: callout-tip
Hint: extract `month` from `Date` with `lubridate::month()`, create `season` with `case_when()` mapping month ranges to labels, then convert to a factor with explicit level ordering:

```{r}
#| eval: false
data |> mutate(
  month = month(Date),
  season = case_when(...),
  season = factor(season, levels = c('Snowmelt', 'Monsoon', 'Low-flow', 'Winter'))
)
```
:::

```{r}
flow_meta <- flow_meta |> 
  mutate(
    month = month(Date),
    season = case_when(
      month %in% 4:6 ~ "Snowmelt",
      month %in% 7:9 ~ "Monsoon",
      month %in% 10:12 ~ "Low-flow",
      month %in% 1:3 ~ "Winter"
    ),
    season = factor(season, levels = c("Snowmelt", "Monsoon", "Low-flow", "Winter"))
  )
```

**b.** Group by gauge, water year, and season. Summarize total seasonal discharge and normalized runoff per group.

```{r}
seasonal_summary <- flow_meta |> 
  group_by(site_no, water_year, season) |> 
  summarize(
    total_season_flow = sum(Flow, na.rm = TRUE),
    normalized_runoff = sum(flow_per_sqmi, na.rm = TRUE),
    .groups = "drop"
  )
```

**c.** For each gauge and water year, identify the **date of peak 7-day mean flow** and convert it to day-of-year. Plot peak day-of-year vs. water year for **Lees Ferry** with a linear trend line.

::: callout-tip
`slice_max(roll7, n = 1, with_ties = FALSE)` inside a `group_by(site_no, water_year)` block selects the single highest 7-day mean row per gauge per year. Then `lubridate::yday(Date)` converts to day-of-year (1–365).
:::

```{r}
peak_flow <- flow_meta |> 
  group_by(site_no, water_year) |> 
  slice_max(roll7, n = 1, with_ties = FALSE) |> 
  mutate(doy = yday(Date)) |> 
  ungroup()
```

```{r}
lees_peak <- peak_flow |> 
  filter(site_no == "09380000")

ggplot(lees_peak, aes(x = water_year, y = doy)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Water Year",
    y = "Peak Flow Day of Year",
    title = "Peak 7-Day Flow Timing at Lees Ferry"
  ) +
  theme_minimal()
```


**d.** In 2–3 sentences: (1) describe the direction and magnitude of the peak timing trend, (2) is it statistically visible or buried in noise? (3) What would a two-week earlier peak timing mean operationally for reservoir managers trying to capture spring runoff before it passes downstream?

> ... The timing of peak 7-day flow at Lees Ferry shows a gradual shift toward earlier dates over the study period, indicating that spring snowmelt runoff is occurring sooner in the year. While the trend is visible in the fitted line, there is considerable year-to-year variability, suggesting the signal is present but somewhat obscured by natural variability in snowpack and climate conditions. If peak runoff occurred two weeks earlier, reservoir managers would need to adjust release schedules and storage strategies to capture snowmelt before it passes downstream, potentially affecting water supply planning and flood control operations.

---

# **Question 9**: Predicting Lees Ferry from Upstream Conditions {#q9}

Lees Ferry is the **Compact's accounting point** — the location where upper basin states must deliver flow to lower basin states. Predicting annual discharge at Lees Ferry from upstream conditions has direct operational value: managers can begin planning deliveries months before the water arrives.

We'll build a simple but physically grounded model: can Glenwood Springs (an upper Rocky Mountain gauge) predict Lees Ferry annual flow, and, does that relationship change by season?

## Data Preparation

**a.** Build your model dataset through four sequential operations:

1. Filter `annual` to Lees Ferry; rename `total_flow` to `lees_flow`
2. Join annual Glenwood Springs flow (site `09085000`); rename to `glenwood_flow`
3. Join the **dominant season** at Lees Ferry per water year, the season with the highest total discharge
4. Log-transform both flow variables

::: callout-tip
The dominant season: within each water year, which season (Snowmelt, Monsoon, Low-flow, Winter) carried the most flow at Lees Ferry? Filter seasonal summaries to Lees Ferry, then `group_by(water_year) |> slice_max(total_flow, n = 1)` to select one season per year.

Then join that back to your Lees Ferry & Glenwood annual table by `water_year`.

This is your fourth join. **Always verify**: the final dataset should have ~44 rows (one per water year, 1980–2023), zero NAs in flow columns, and log-transformed values should be positive.
:::

```{r}
lees_model_data <- annual_summary |>
  
  # 1. Filter to Lees Ferry
  filter(site_no == "09380000") |>
  select(water_year, total_flow) |>
  rename(lees_flow = total_flow) |>
  
  # 2. Join Glenwood Springs flow
  left_join(
    annual_summary |>
      filter(site_no == "09085000") |>
      select(water_year, total_flow) |>
      rename(glenwood_flow = total_flow),
    by = "water_year"
  ) |>
  
  # 3. Join dominant season at Lees Ferry
  left_join(
    seasonal_summary |>
      filter(site_no == "09380000") |>
      group_by(water_year) |>
      slice_max(total_season_flow, n = 1, with_ties = FALSE) |>
      ungroup() |>
      select(water_year, season),
    by = "water_year"
  ) |>
  
  # 4. Log transform flows
  mutate(
    log_lees = log(lees_flow),
    log_glenwood = log(glenwood_flow)
  )
```

```{r}
nrow(lees_model_data)
```

```{r}
summary(lees_model_data)
```


## Model Building

**b.** Fit a linear model predicting `log_lees` from `log_glenwood`, `season`, and their interaction, If you haven't seen linear modeling in R, we will cover it in more detail latter in this course, but, the syntax is straightforward: `lm(response ~ predictor1 * predictor2, data = mydata)` fits a model with main effects and an interaction term. The interaction (`*`) allows the relationship between Glenwood and Lees to differ by season while a model without interaction (`lm(log_lees ~ log_glenwood + season)`) would assume the same Glenwood-Lees relationship in every season.

```{r}
mod <- lm(log_lees ~ log_glenwood * season, data = lees_model_data)
summary(mod)
```


**c.** In 3–4 sentences: (1) report the R² and state whether the overall model is statistically significant, (2) interpret the `log_glenwood` coefficient as an elasticity — what does it mean physically that the coefficient is greater than 1? (3) describe what the significant negative interaction terms tell you about the upstream-downstream relationship outside of Snowmelt season.

::: callout-tip
In a log-log model, the coefficient on `log_glenwood` is an **elasticity**: a coefficient of 1.8 means a 1% increase in Glenwood Springs flow is associated with a 1.8% increase at Lees Ferry. This super-linear amplification makes physical sense — flow accumulates from tributaries between the two gauges, so an above-average headwater year compounds as it moves downstream.
:::

> ... The model explains a large portion of the variability in Lees Ferry annual flow, with an R² of 0.855 and a highly significant overall F-test (p ≈ 1.12 × 10⁻¹³), indicating that Glenwood Springs flow and season jointly predict downstream flow well. The log_glenwood coefficient of 1.797 represents an elasticity: a 1% increase in Glenwood flow corresponds to roughly a 1.8% increase at Lees Ferry, reflecting super-linear accumulation from tributaries between the two gauges. The significant negative interaction terms for Monsoon, Low-flow, and Winter indicate that the upstream-downstream relationship is weaker outside of Snowmelt, meaning the amplification of flow is smaller in these seasons compared to Snowmelt. In other words, Glenwood flow has the largest effect on Lees Ferry during Snowmelt, while seasonal conditions modulate this relationship for the rest of the year.

---

# **Question 10**: Model Evaluation {#q10}

A model that fits well on average can still fail in predictable ways. Two core assumptions of linear regression are: predictions should be unbiased across the full range of fitted values, and residuals should be approximately normally distributed. Let's check both — and look for the outlier year that the model cannot predict.

**a.** Use `broom::augment()` to produce a tidy data frame of predictions and residuals alongside the original data:

```{r}
#| eval: false
aug <- augment(mod, data = lees_model_data)
```

::: callout-tip
## Introducing `broom`

`broom` converts messy R model objects into tidy, pipeable data frames:

- `augment(mod)` — adds `.fitted` (predictions) and `.resid` (residuals) to your data
- `glance(mod)` — one-row summary of model fit: R², AIC, p-value
- `tidy(mod)` — one-row-per-coefficient table with estimates and p-values

Together these make model results as pipeable as any other data frame — straight into `ggplot` or `flextable`.
:::

```{r}

aug <- augment(mod, data = lees_model_data)
head(aug)
```

**b.** Create a **predicted vs. actual scatter plot**. Add a dashed 1:1 reference line (perfect prediction). Color points by `season`. Identify any point that sits noticeably off the line.

::: callout-tip
`geom_abline(slope = 1, intercept = 0, linetype = "dashed")` draws the 1:1 line. Points above it = underpredicted (actual > predicted); points below = overpredicted.
:::

```{r}

ggplot(aug, aes(x = .fitted, y = log_lees, color = season)) +
  geom_point(size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Predicted log(Lees Ferry flow)", y = "Observed log(Lees Ferry flow)",
       title = "Predicted vs. Actual Lees Ferry Flow by Season") +
  theme_minimal()
```

**c.** Plot a **histogram of the residuals** with a vertical dashed line at zero.

```{r}
ggplot(aug, aes(x = .resid)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Residuals", y = "Count", title = "Histogram of Model Residuals") +
  theme_minimal()
```

**d.** In 3–4 sentences: (1) identify the outlier water year visible in your scatter plot. Which year is it and why did the model fail to predict it? (2) Do the residuals appear approximately normal, or is there a tail that concerns you? (3) Name two additional predictors that would most improve this model and explain the physical mechanism behind each.

> ... The scatter plot reveals an outlier in water year 1983, which the model substantially underpredicted. This year likely experienced extreme flow events not captured by upstream Glenwood Springs discharge alone, such as anomalously high tributary inflows or a wet Monsoon season. The residual histogram is roughly normal but shows a slight positive tail, indicating occasional underprediction of very high flows. Including precipitation anomalies in the Lees Ferry watershed and snowpack storage upstream as additional predictors would likely improve the model. Precipitation directly adds to flow, while snowpack determines delayed melt contributions, both capturing physical processes beyond Glenwood flow.

---

# Going Further (ungraded)

The model in Q9 predicts Lees Ferry from a single upstream gauge. `dataRetrieval` gives you access to hundreds more across the basin. Could a multi-gauge ensemble prediction outperform the single-gauge model? Could peak flows in March predict the following June's flow at Lees Ferry months before snowmelt begins?

These are real operational forecasting questions that USBR and state water agencies work on every year. The data is already in your hands to try!

---

# Summary

In this lab you built a complete analytical pipeline for one of the most consequential river systems in North America:

- **Retrieved** 44 years of daily streamflow directly from the USGS NWIS API using `dataRetrieval`
- **Joined** gauge metadata using relational keys — and verified every join
- **Normalized** by drainage area to enable fair comparisons across the basin
- **Detected** low-flow threshold exceedances using rolling statistics for operational drought monitoring
- **Aggregated** by water year and season — the natural temporal units of western hydrology
- **Visualized** basin-wide trends with a standardized anomaly heatmap and a flow-weighted spatial center analysis
- **Modeled** the upstream-downstream relationship at the Compact's accounting point
- **Evaluated** model assumptions and identified the outlier year the model couldn't predict

Every tool — `dplyr`, `ggplot2`, `zoo`, `patchwork`, `broom`, `flextable` — recurs in work you will inevitabbly do. The key is that the workflow is always the same: get → tidy → explore → model → communicate.

---

# Rubric

| Question | Topic | Points |
|----------|-------|--------|
| Q1 | Daily conditions report | 10 |
| Q2 | Basin metadata join | 10 |
| Q3 | Per-unit-area normalization | 10 |
| Q4 | Low-flow threshold monitoring | 20 |
| Q5 | Annual water year summary | 10 |
| Q6 | Basin-wide trend analysis | 15 |
| Q7 | Moving center of drought | 15 |
| Q8 | Seasonal regimes + peak timing | 15 |
| Q9 | Predictive model | 15 |
| Q10 | Model evaluation | 10 |
| Well-structured, legible Qmd | — | 10 |
| Deployed as a webpage | — | 10 |
| **Total** | | **150** |

---

**Grading key (brief):**
- **Correctness (primary):** calculations, joins, and checks — e.g., Q2 join verification and Q4 rolling mean logic.
- **Reproducibility:** code runs end-to-end, used parameter objects (`my.date`), and included any small checks or assertions.
- **Clarity & Presentation:** readable code, labeled tables/plots, captions for `flextable` outputs, and a clear 1-paragraph interpretation.

# Submission

1. Render: `quarto render lab-01.qmd`
2. Push: `git add -A && git commit -m "lab 1 complete" && git push`
3. Activate GitHub Pages: **GitHub → Settings → Pages → Deploy from branch → main**
4. Submit your live URL to Canvas: `https://USERNAME.github.io/csu-523c/lab-01.html`

Add this project to your personal website with a brief description — it counts toward end-of-semester extra credit.

```{r}
rmarkdown::render("lab-01.qmd", output_format = "html_document", output_dir = "docs")
```

